---
title: "MultiNicheNet analysis MIS-C: M vs S vs A"
author: "Robin Browaeys"
date: "2022-05-05"
---
```{r}
library(tidyverse)
library(qsub)
```

```{r}
sce_path = "/group/irc/personal/robinb/MIS-C/data/sce_misc.rds"
output_path = "/group/irc/personal/robinb/MIS-C/"
sample_id = "ShortID"
group_id = "MIS.C.AgeTier"
celltype_id = "Annotation_v2.0"
covariates = NA 
batches = NA 
min_cells = 10
```

```{r}
contrasts_oi = c("'M-(S+A)/2','S-(M+A)/2','A-(S+M)/2'")
contrast_tbl = tibble(contrast = 
                        c("M-(S+A)/2","S-(M+A)/2", "A-(S+M)/2"), 
                      group = c("M","S","A"))
```

```{r}
empirical_pval = FALSE
p_val_adj = FALSE 
logFC_threshold = 0.50
p_val_threshold = 0.05
fraction_cutoff = 0.05
top_n_target = 250
```

```{r}
prioritizing_weights_DE = c("de_ligand" = 1,
                         "de_receptor" = 1)
prioritizing_weights_activity = c("activity_scaled" = 2)

prioritizing_weights_expression_specificity = c("exprs_ligand" = 2,
                         "exprs_receptor" = 2)

prioritizing_weights_expression_sufficiency = c("frac_exprs_ligand_receptor" = 1)

prioritizing_weights_relative_abundance = c( "abund_sender" = 0,
                         "abund_receiver" = 0)

```

```{r}
prioritizing_weights = c(prioritizing_weights_DE, 
                         prioritizing_weights_activity, 
                         prioritizing_weights_expression_specificity,
                         prioritizing_weights_expression_sufficiency, 
                         prioritizing_weights_relative_abundance)
```


```{r}
multinichenet_wrapper = function(i, sce_path, output_path, celltype_id, sample_id, group_id, contrasts_oi, contrast_tbl, batches, covariates, min_cells, prioritizing_weights, logFC_threshold, p_val_threshold, fraction_cutoff, p_val_adj, empirical_pval, top_n_target){
  
  library(SingleCellExperiment)
  library(tidyverse)
  library(nichenetr)
  library(multinichenetr)
  
  ### read in NicheNet model
  lr_network = readRDS("/group/irc/personal/robinb/lr_network_human_21122021.rds")
  lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor) %>% mutate(ligand = make.names(ligand), receptor = make.names(receptor))
  ligand_target_matrix = readRDS("/group/irc/personal/robinb/ligand_target_matrix_nsga2r_final.rds")
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% make.names()
  
  ### read in scRNAseq data
  sce = readRDS(sce_path)
  print(SummarizedExperiment::colData(sce) %>% head())
  sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()

  ### DE analysis

  DE_info = get_DE_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, batches = batches, covariates = covariates, contrasts_oi = contrasts_oi, min_cells = min_cells)

  print("DE_info")
  
  markobj = c('---',
             'title: "P-value Histograms"',
             'output: html_document',
             '---',
             '',
             '## hist_pvals',
             '',
             '```{r, fig.width = 21, fig.height = 9}',
             'DE_info$hist_pvals',
             '```',
             ''
             )

  markdown::markdownToHTML(text = knitr::knit(text = markobj), output = paste0(output_path, "plots/hist_pvals.html"))

  DE_info$hist_pvals = NULL

  if(empirical_pval == TRUE){
    DE_info_emp = get_empirical_pvals(DE_info$celltype_de$de_output_tidy)
    comparison_plots = compare_normal_emp_pvals(DE_info, DE_info_emp, adj_pval = FALSE)
    celltype_de = DE_info_emp$de_output_tidy_emp %>% dplyr::select(-p_val, -p_adj) %>% dplyr::rename(p_val = p_emp, p_adj = p_adj_emp)
    markobj = c('---',
             'title: "Empericial p-value Plots"',
             'output: html_document',
             '---',
             '',
             '## comparison_plots',
             '',
             '```{r, fig.width = 10, fig.height = 10}',
             'comparison_plots',
             '```',
              '',
             '## hist_pvals_emp',
             '',
             '```{r, fig.width = 21, fig.height = 9}',
             'DE_info_emp$hist_pvals_emp',
             '```',
             '',
             '## z_distr_plots_emp_pval',
             '',
             '```{r, fig.width = 10, fig.height = 10}',
             'DE_info_emp$z_distr_plots_emp_pval',
             '```',
             ''
             )

    markdown::markdownToHTML(text = knitr::knit(text = markobj), output = paste0(output_path, "plots/emp_DE_plots.html"))
  } else {
    celltype_de = DE_info$celltype_de$de_output_tidy
  }
  
  senders_oi = celltype_de$cluster_id %>% unique()
  receivers_oi = celltype_de$cluster_id %>% unique()
  genes_oi = celltype_de$gene %>% unique()
  sce = sce[genes_oi, SummarizedExperiment::colData(sce)[,celltype_id] %in% c(senders_oi, receivers_oi)]

  sender_receiver_de = combine_sender_receiver_de(
    sender_de = celltype_de,
    receiver_de = celltype_de,
    senders_oi = senders_oi,
    receivers_oi = receivers_oi,
    lr_network = lr_network
  )
  sender_receiver_tbl = sender_receiver_de %>% dplyr::distinct(sender, receiver)
  
    ### abundance_expression_info

  abundance_expression_info = get_abundance_expression_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, min_cells = min_cells, senders_oi = senders_oi, receivers_oi = receivers_oi, lr_network = lr_network, batches = batches)

  markobj = c('---',
             'title: "Abundance Plots"',
             'output: html_document',
             '---',
             '',
             '## abund_plot_sample',
             '',
             '```{r, fig.width = 12, fig.height = 25}',
             'abundance_expression_info$abund_plot_sample',
             '```',
              '',
             '## abund_plot_group',
             '',
             '```{r, fig.width = 12, fig.height = 12}',
             'abundance_expression_info$abund_plot_group',
             '```',
             '',
             '## abund_barplot',
             '',
             '```{r, fig.width = 12, fig.height = 12}',
             'abundance_expression_info$abund_barplot',
             '```',
             ''
             )

  markdown::markdownToHTML(text = knitr::knit(text = markobj), output = paste0(output_path, "plots/abundance_plots.html"))
  
  abundance_expression_info$abund_plot_sample = NULL
  abundance_expression_info$abund_plot_group = NULL
  abundance_expression_info$abund_barplot = NULL
  
  metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

  if(!is.na(batches)){
    grouping_tbl = metadata_combined[,c(sample_id, group_id, batches)] %>% tibble::as_tibble() %>% dplyr::distinct()
    colnames(grouping_tbl) = c("sample","group",batches)
  } else {
    grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% tibble::as_tibble() %>% dplyr::distinct()
    colnames(grouping_tbl) = c("sample","group")
  }

  
  rm(sce)
  
  ### ligand activities
  
  n.cores = min(3, length(receivers_oi))

  ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(get_ligand_activities_targets_DEgenes(
    receiver_de = celltype_de,
    receivers_oi = receivers_oi,
    ligand_target_matrix = ligand_target_matrix,
    logFC_threshold = logFC_threshold,
    p_val_threshold = p_val_threshold,
    p_val_adj = p_val_adj,
    top_n_target = top_n_target,
    verbose = TRUE, 
    n.cores = n.cores
  )))
  
  ### save intermediary output
  # list(abundance_expression_info = abundance_expression_info, grouping_tbl = grouping_tbl, sender_receiver_de = sender_receiver_de, ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes, contrast_tbl = contrast_tbl, sender_receiver_tbl = sender_receiver_tbl, celltype_de = celltype_de) %>% saveRDS(paste0(output_path,"output/MNN_intermediary_output.rds"))
  


  ### Prioritization tables
  prioritization_tables = suppressMessages(generate_prioritization_tables(
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    contrast_tbl = contrast_tbl,
    sender_receiver_tbl = sender_receiver_tbl,
    grouping_tbl = grouping_tbl,
    prioritizing_weights = prioritizing_weights,
    fraction_cutoff = fraction_cutoff, 
    abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
    abundance_data_sender = abundance_expression_info$abundance_data_sender
))
  
  ## LR prod matrix
  return_lr_prod_matrix = FALSE
if(return_lr_prod_matrix == TRUE){
    
    ids_oi = prioritization_tables$group_prioritization_tbl %>% dplyr::filter(fraction_expressing_ligand_receptor > 0)  %>% dplyr::pull(id) %>% unique()
    
    lr_prod_df = abundance_expression_info$sender_receiver_info$pb_df %>% dplyr::inner_join(grouping_tbl, by = "sample") %>% dplyr::mutate(lr_interaction = paste(ligand, receptor, sep = "_")) %>% dplyr::mutate(id = paste(lr_interaction, sender, receiver, sep = "_")) %>% dplyr::select(sample, id, ligand_receptor_pb_prod) %>% dplyr::filter(id %in% ids_oi) %>% dplyr::distinct() %>% tidyr::spread(id, ligand_receptor_pb_prod)
    lr_prod_mat = lr_prod_df %>% dplyr::select(-sample) %>% data.frame() %>% as.matrix()
    rownames(lr_prod_mat) = lr_prod_df$sample
    
    col_remove = lr_prod_mat %>% apply(2,function(x)sum(x != 0)) %>% .[. == 0] %>% names()
    row_remove = lr_prod_mat %>% apply(1,function(x)sum(x != 0)) %>% .[. == 0] %>% names()
    
    lr_prod_mat = lr_prod_mat %>% .[rownames(.) %>% generics::setdiff(col_remove),colnames(.) %>% generics::setdiff(col_remove)]
  } else {
    lr_prod_mat = NULL
}

  ## correlation
  
  lr_target_prior_cor = lr_target_prior_cor_inference(prioritization_tables$group_prioritization_tbl$receiver %>% unique(), abundance_expression_info, celltype_de, grouping_tbl, prioritization_tables, ligand_target_matrix, logFC_threshold = logFC_threshold, p_val_threshold = p_val_threshold, p_val_adj = p_val_adj)

  ## save output
  multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    # sender_receiver_info = abundance_expression_info$sender_receiver_info,
    # sender_receiver_de =  sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    lr_prod_mat = lr_prod_mat,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  ) 
  multinichenet_output = make_lite_output(multinichenet_output)
  
  saveRDS(multinichenet_output, paste0(output_path,"output/MNN_output_MSA.rds"))

  return(multinichenet_output)

}

```

```{r}
qsub_config = create_qsub_config(
  remote = "robinb@prism.psb.ugent.be:7777",
  local_tmp_path = "/home/robin/r2gridengine",
  remote_tmp_path = "/scratch/irc/personal/robinb/r2gridengine",
  modules = "R/x86_64/4.0.3",
  memory = "240G",
  wait = FALSE,
  remove_tmp_folder = FALSE,
  name = "MNN-MISC",
  max_wall_time = "500:00:00",
  stop_on_error = TRUE,
  num_cores = 1
)
job_MNN = qsub_lapply(X = 1, FUN = multinichenet_wrapper,
                           object_envir = environment(multinichenet_wrapper),
                           qsub_config = qsub_config,
                           qsub_environment = NULL,
                           qsub_packages = NULL, 
                            sce_path, output_path, celltype_id, sample_id, group_id, contrasts_oi, contrast_tbl, batches, covariates, min_cells, prioritizing_weights, logFC_threshold, p_val_threshold, fraction_cutoff, p_val_adj, empirical_pval, top_n_target)
saveRDS(job_MNN, "../output/job_MNN.rds")

# output = qsub_retrieve(job_MNN)
```


```{bash}
scp -r robinb@prism:/group/irc/personal/robinb/MIS-C/plots/* /media/sf_work/Research/NicheNet/MultiNicheNet\ paper/code/case_studies/MIS-C/plots_multinichenet/
```

```{bash}
scp -r robinb@prism:/group/irc/personal/robinb/MIS-C/output/MNN_output_MSA.rds /media/sf_work/Research/NicheNet/MultiNicheNet\ paper/code/case_studies/MIS-C/output/
```




