---
title: "MultiNicheNet analysis: MIS-C threewise comparison - step-by-step"
author: "Robin Browaeys"
date: "2023-06-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MultiNicheNet analysis: MIS-C threewise comparison - step-by-step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- github markdown built using 
rmarkdown::render("vignettes/basic_analysis_steps_MISC.Rmd", output_format = "github_document")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

In this vignette, you can learn how to perform an all-vs-all MultiNicheNet analysis. In this vignette, we start from one SingleCellExperiment object containing cells from both sender and receiver cell types and from different patients.

A MultiNicheNet analysis can be performed if you have multi-sample, multi-group single-cell data. MultiNicheNet will look for cell-cell communication between the cell types in your data for each sample, and compare the cell-cell communication patterns between the groups of interest. Therefore, the absolute minimum of meta data you need to have, are following columns indicating for each cell: the **group**, **sample** and **cell type**.
 
As example expression data of interacting cells, we will here use scRNAseq data of immune cells in MIS-C patients and healthy siblings from this paper of Hoste et al.: [TIM3+ TRBV11-2 T cells and IFNγ signature in patrolling monocytes and CD16+ NK cells delineate MIS-C](https://rupress.org/jem/article/219/2/e20211381/212918/TIM3-TRBV11-2-T-cells-and-IFN-signature-in) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6362434.svg)](https://doi.org/10.5281/zenodo.6362434)
. MIS-C (multisystem inflammatory syndrome in children) is a novel rare immunodysregulation syndrome that can arise after SARS-CoV-2 infection in children. We will use NicheNet to explore immune cell crosstalk enriched in MIS-C compared to healthy siblings. 

In this vignette, we will prepare the data and analysis parameters, and then perform the MultiNicheNet analysis. 

The different steps of the MultiNicheNet analysis are the following:

* 0. Preparation of the analysis: load packages, NicheNet LR network & ligand-target matrix, single-cell expression data, and define main settings of the MultiNicheNet analysis

* 1. Extract cell type abundance and expression information from receiver and sender cell types, and link this expression information for ligands of the sender cell types to the corresponding receptors of the receiver cell types 

* 2. Perform genome-wide differential expression analysis of receiver and sender cell types to define DE genes between the conditions of interest. Based on this analysis, we can define the logFC/p-value of ligands in senders and receptors in receivers, and define the set of affected target genes in the receiver.

* 3. Predict NicheNet ligand activities and NicheNet ligand-target links based on these differential expression results

* 4. Use the information collected above to prioritize all sender-ligand---receiver-receptor pairs.

* 5. Calculate correlation in expression between ligand-receptor pairs and their predicted target genes

In this vignette, we will demonstrate all these steps one by one.

After the MultiNicheNet analysis is done, we will explore the output of the analysis with different ways of visualization. 

# Step 0: Preparation of the analysis: load packages, NicheNet LR network & ligand-target matrix, single-cell expression data

## Step 0.1: Load required packages and NicheNet ligand-receptor network and ligand-target matrix

```{r}
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(multinichenetr)
```

The Nichenet v2 networks and matrices for both mouse and human can be downloaded from Zenodo [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.7074291.svg)](https://doi.org/10.5281/zenodo.7074291). 

We will read these object in for human because our expression data is of human patients. 
Gene names are here made syntactically valid via `make.names()` to avoid the loss of genes (eg H2-M3) in downstream visualizations.

```{r}
organism = "human"
if(organism == "human"){
  lr_network = readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
  lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor) %>% mutate(ligand = make.names(ligand), receptor = make.names(receptor))
  ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% make.names()
} else if(organism == "mouse"){
  lr_network = readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
  lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor) %>% mutate(ligand = make.names(ligand), receptor = make.names(receptor))
  ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% make.names()
}
```

## Step 0.2: Read in SingleCellExperiment Objects 

In this vignette, sender and receiver cell types are in the same SingleCellExperiment object, which we will load here. In this vignette, we will load in a subset of the scRNAseq data of the MIS-C [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.8010790.svg)](https://doi.org/10.5281/zenodo.8010790). For the sake of demonstration, this subset only contains 3 cell types. These celltypes are some of the cell types that were found to be most interesting related to MIS-C according to Hoste et al. 

If you start from a Seurat object, you can convert it easily to a SingleCellExperiment via `sce = Seurat::as.SingleCellExperiment(seurat_obj, assay = "RNA")`.

Because the NicheNet 2.0. networks are in the most recent version of the official gene symbols, we will make sure that the gene symbols used in the expression data are also updated (= converted from their "aliases" to official gene symbols). Afterwards, we will make them again syntactically valid. 

```{r}
sce = readRDS(url("https://zenodo.org/record/8010790/files/sce_subset_misc.rds"))
sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()
```

## Step 0.3: Prepare settings of the MultiNicheNet cell-cell communication analysis

### Define in which metadata columns we can find the **group**, **sample** and **cell type** IDs

In this case study, we want to study differences in cell-cell communication patterns between MIS-C patients (M), their healthy siblings (S). The meta data columns that indicate this disease status is `MIS.C.AgeTier`. This datasets also contains cells from adult patients with severe COVID (A), however we are not interested in this comparison in this vignette because the additional data modality (proteomics) we have only covers MIS-C and healthy siblings.

Cell type annotations are indicated in the `Annotation_v2.0` column, and the sample is indicated by the `ShortID` column. 
If your cells are annotated in multiple hierarchical levels, we recommend using a high level in the hierarchy. This for 2 reasons: 1) MultiNicheNet focuses on differential expression and not differential abundance, and 2) there should be sufficient cells per sample-celltype combination.

If you would have batch effects or covariates you can correct for, you can define this here as well. 

Important: for categorical covariates and batches, there should be at least one sample for every group-batch combination. If one of your groups/conditions lacks a certain level of your batch, you won't be able to correct for the batch effect because the model is then not able to distinguish batch from group/condition effects.

Important: the column names of group, sample, cell type, batches and covariates should be syntactically valid (`make.names`)

Important: All group, sample, cell type, batch and covariate names should be syntactically valid as well (`make.names`) (eg through `SummarizedExperiment::colData(sce)$ShortID = SummarizedExperiment::colData(sce)$ShortID %>% make.names()`)

```{r}
sample_id = "ShortID"
group_id = "MIS.C.AgeTier"
celltype_id = "Annotation_v2.0"
covariates = NA
batches = NA
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% c("M","S")]
```

Sender and receiver cell types also need to be defined. Both are here all cell types in the dataset because we are interested in an All-vs-All analysis.

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
```

If the user wants it, it is possible to use only a subset of senders and receivers. Senders and receivers can be entirely different, but also overlapping, or the same. If you don't use all the cell types in your data, we recommend to continue with a subset of your data.

```{r}
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% c(senders_oi, receivers_oi)]
```

In this vignette we will focus on comparing the MI
Now we will go to the first real step of the MultiNicheNet analysis

# Step 1: Extract cell type abundance of receiver and sender cell types

Since MultiNicheNet will infer group differences at the sample level for each cell type (currently via Muscat - pseudobulking + EdgeR), we need to have sufficient cells per sample of a cell type, and this for both groups. In the following analysis we will set this minimum number of cells per cell type per sample at 10 (recommended minimum). 

```{r}
min_cells = 10
```

Now we will calculate abundance and expression information for each cell type / sample / group combination with the following functions. In the output of this function, you can also find some 'Cell type abundance diagnostic plots' that will the users which celltype-sample combinations will be left out later on for DE calculation because the nr of cells is lower than de defined minimum defined here above. If too many celltype-sample combinations don't pass this threshold, we recommend to define your cell types in a more general way (use one level higher of the cell type ontology hierarchy) (eg TH17 CD4T cells --> CD4T cells).

```{r}
abundance_info = make_abundance_plots(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, min_cells = min_cells, senders_oi = senders_oi, receivers_oi = receivers_oi, batches = batches)
```

First, we will check the cell type abundance diagnostic plots.

### Interpretation of cell type abundance information

The first plot visualizes the number of cells per celltype-sample combination, and indicates which combinations are removed during the DE analysis because there are less than `min_cells` in the celltype-sample combination. 

```{r}
abundance_info$abund_plot_sample
```
The red dotted line indicates the required minimum of cells as defined above in `min_cells`. We can see here that some sample-celltype combinations are left out. For the DE analysis in the next step, only cell types will be considered if there are at least two samples per group with a sufficient number of cells. For these cell types in this dataset, all cell types are retained.

```{r}
abundance_info$abund_plot_group
```

```{r}
abundance_info$abund_barplot
```
__Important__: Based on the cell type abundance diagnostics, we recommend users to change their analysis settings if required (eg changing cell type annotation level, batches, ...), before proceeding with the rest of the analysis.

# Step 2: Perform genome-wide differential expression analysis of receiver and sender cell types to define DE genes between the conditions of interest. Based on this analysis, we can define the logFC/p-value of ligands in senders and receptors in receivers, and define the set of affected target genes in the receiver.

Now we will go over to the multi-group, multi-sample differential expression (DE) analysis (also called 'differential state' analysis by the developers of Muscat).

### Define the contrasts and covariates of interest for the DE analysis.

Here, we want to compare each patient group to the other groups, so the MIS-C (M) group vs healthy control siblins (S) and adult COVID19 patients (A) etcetera. 
To do this comparison, we need to set the following contrasts:

```{r}
contrasts_oi = c("'M-S','S-M'")
```

__Very Important__ Note the format to indicate the contrasts! This formatting should be adhered to very strictly, and white spaces are not allowed!  Check `?get_DE_info` for explanation about how to define this well. The most important things are that: each contrast is surrounded by single quotation marks, contrasts are separated by a comma without any whitespace, and alle contrasts together are surrounded by double quotation marks. If you compare against two groups, you should divide by 2, if you compare against three groups, you should divide by 3 etcetera.

For downstream visualizations and linking contrasts to their main group, you need to run the following:

```{r}
contrast_tbl = tibble(contrast = 
                        c("M-S","S-M"), 
                      group = c("M","S"))
```


### Perform the DE analysis for each cell type.

```{r}
DE_info = get_DE_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, batches = batches, covariates = covariates, contrasts_oi = contrasts_oi, min_cells = min_cells)
```

### Check DE results

Table with logFC and p-values for each gene-celltype-contrast:

```{r}
DE_info$celltype_de$de_output_tidy %>% arrange(p_adj) %>% head()
```

```{r}
celltype_de = DE_info$celltype_de$de_output_tidy
```

Only keep cell types in the SingleCellExperiment object if they were retained in the DE analysis
```{r}
retained_celltypes = celltype_de$cluster_id %>% unique()
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% retained_celltypes]
```

In the next step, we will combine the DE information of senders and receivers by linking their ligands and receptors together:

### Combine DE information for ligand-senders and receptors-receivers (similar to step1 - `abundance_expression_info$sender_receiver_info`)

```{r}
sender_receiver_de = combine_sender_receiver_de(
  sender_de = celltype_de,
  receiver_de = celltype_de,
  senders_oi = senders_oi,
  receivers_oi = receivers_oi,
  lr_network = lr_network
)
```

```{r}
sender_receiver_de %>% head(20)
```
### Prepare sender-receiver and sample-condition grouping tables necessary for downstream analyses

```{r}
sender_receiver_tbl = sender_receiver_de %>% dplyr::distinct(sender, receiver)

metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

if(!is.na(batches)){
  grouping_tbl = metadata_combined[,c(sample_id, group_id, batches)] %>% tibble::as_tibble() %>% dplyr::distinct()
  colnames(grouping_tbl) = c("sample","group",batches)
} else {
  grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% tibble::as_tibble() %>% dplyr::distinct()
  colnames(grouping_tbl) = c("sample","group")
}
```

Crucial note: grouping_tbl: group should be the same as in the contrast_tbl, and as in the expression info tables! Rename accordingly if this would not be the case. If you followed the guidelines of this tutorial closely, there should be no problem.

# Step 3: Extract expression information from receiver and sender cell types, and link this expression information for ligands of the sender cell types to the corresponding receptors of the receiver cell types

```{r}
abundance_expression_info = get_abundance_expression_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, min_cells = min_cells, senders_oi = senders_oi, receivers_oi = receivers_oi, lr_network = lr_network, batches = batches)
```

# Step 4: Predict NicheNet ligand activities and NicheNet ligand-target links based on these differential expression results

## Define the parameters for the NicheNet ligand activity analysis 

Here, we need to define the thresholds that will be used to consider genes as differentially expressed or not (logFC, p-value, decision whether to use adjusted or normal p-value, minimum fraction of cells that should express a gene in at least one sample in a group, whether to use the normal p-values or empirical p-values). 

NicheNet ligand activity will then be calculated as the enrichment of predicted target genes of ligands in this set of DE genes compared to the genomic background. Here we choose for a minimum logFC of 0.50, maximum p-value of 0.05, and minimum fraction of expression of 0.05. 

```{r}
logFC_threshold = 0.50
p_val_threshold = 0.05
fraction_cutoff = 0.05
```

We will here choose for applying the p-value cutoff on the normal p-values, and not on the p-values corrected for multiple testing. This choice was made here because this dataset has only a few samples per group and we might have a lack of statistical power due to pseudobulking. In case of more samples per group, and a sufficient high number of DE genes per group-celltype (> 50), we would recommend using the adjusted p-values.

```{r}
# p_val_adj = TRUE 
p_val_adj = FALSE 
```

For the NicheNet ligand-target inference, we also need to select which top n of the predicted target genes will be considered (here: top 250 targets per ligand).

```{r}
top_n_target = 250
```

The NicheNet ligand activity analysis can be run in parallel for each receiver cell type, by changing the number of cores as defined here. This is only recommended if you have many receiver cell type. 

```{r}
verbose = TRUE
cores_system = 8
n.cores = min(cores_system, sender_receiver_de$receiver %>% unique() %>% length()) # use one core per receiver cell type
```

## Run the NicheNet ligand activity analysis 

(this might take some time)
```{r}
ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(get_ligand_activities_targets_DEgenes(
  receiver_de = celltype_de,
  receivers_oi = receivers_oi,
  ligand_target_matrix = ligand_target_matrix,
  logFC_threshold = logFC_threshold,
  p_val_threshold = p_val_threshold,
  p_val_adj = p_val_adj,
  top_n_target = top_n_target,
  verbose = verbose, 
  n.cores = n.cores
)))
```

Check the DE genes used for the activity analysis

```{r}
ligand_activities_targets_DEgenes$de_genes_df %>% head(20)
```

Check the output of the activity analysis

```{r}
ligand_activities_targets_DEgenes$ligand_activities %>% head(20)
```

# Step 5: Use the information collected above to prioritize all sender-ligand---receiver-receptor pairs.

In the previous steps, we calculated expression, differential expression and NicheNet activity information. Now we will combine these different types of information in one prioritization scheme.

MultiNicheNet uses the following criteria to prioritize ligand-receptor interactions:

* Upregulation of the ligand in a sender cell type and/or upregulation of the receptor in a receiver cell type - in the condition of interest. : `de_ligand` and `de_receptor`
* Sufficiently high expression levels of ligand and receptor in many samples of the same group (to mitigate the influence of outlier samples). : `frac_exprs_ligand_receptor`
* Cell-type and condition specific expression of the ligand in the sender cell type and receptor in the receiver cell type (to mitigate the influence of upregulated but still relatively weakly expressed ligands/receptors) : `exprs_ligand` and `exprs_receptor`
* High NicheNet ligand activity, to further prioritize ligand-receptor pairs based on their predicted effect of the ligand-receptor interaction on the gene expression in the receiver cell type : `activity_scaled` 

## Run the prioritization 

```{r}
prioritization_tables = suppressMessages(generate_prioritization_tables(
  sender_receiver_info = abundance_expression_info$sender_receiver_info,
  sender_receiver_de = sender_receiver_de,
  ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
  contrast_tbl = contrast_tbl,
  sender_receiver_tbl = sender_receiver_tbl,
  grouping_tbl = grouping_tbl,
  scenario = "regular", # all prioritization criteria will be weighted equally
  fraction_cutoff = fraction_cutoff, 
  abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
  abundance_data_sender = abundance_expression_info$abundance_data_sender,
  ligand_activity_down = FALSE # use only upregulatory ligand activities to prioritize
))
```

Check the output tables

First: group-based summary table

```{r}
prioritization_tables$group_prioritization_tbl %>% head(20)
```

# Step 6: Add additional data modality information to the prioritization table 

Here we will demonstrate how to tailor the prioritization to additional data modalities that you may have. How to do this exactly will stronglly depend on the type of data modality. Here we should an example we recommend for integrating "targeted" serum proteomics data.

As additional data modality, we will use serum proteomics data (generated through the OLINK platform) that was published by a different research group based on serum samples of a different cohort of MIS-C patients. 
Ligands that are also upregulated in the serum at the protein level may be interesting candidates for further follow-up. Therefore, we can use this data to help in the prioritization. We will here add an additional score to each interaction related to how strongly the ligand is differentially expressed at the protein level in the serum of patients. 

Because we don't have OLINK information of each ligand, we will consider ligands without data as being not-differentially expressed at all (p-val = 1, logFC = 0). By doing this, these ligands will still be kept in the total prioritization analysis such that they could still be ranked high if the previous criteria strongly point to their importance. 

Whereas we will only consider ligand upregulation for prioritization, we will still keep receptor information in the OLINK dataframe for later visualizations.

## Read in additional data modality - here: OLINK serum proteomics data -- keep only ligands and receptors

```{r}
olink_df = xlsx::read.xlsx("/Users/robinb/Work/current_projects/MISC OLINK/data/summary_difference_diorioMIS-C-vs-HC.xlsx", 1) %>% as_tibble() %>% dplyr::rename(gene = variables, logFC = mean_diff_d0, pval = adj_p_values) %>% dplyr::select(gene, logFC, pval) %>% dplyr::mutate(gene = gene %>% make.names()) %>% mutate(olink_type = "Diorio")

olink_df %>% filter(gene %in% lr_network$ligand) %>% nrow() # nr of ligands in OLINK data
olink_df %>% filter(gene %in% lr_network$receptor) %>% nrow() # nr of receptors in OLINK data

olink_df_ligands_receptors = olink_df %>% filter(gene %in% union(lr_network$ligand, lr_network$receptor))
olink_df_ligands_receptors_NAs = tibble(gene = union(lr_network$ligand, lr_network$receptor) %>% setdiff(olink_df_ligands_receptors$gene), logFC = 0, pval = 1) 
olink_df_ligands_receptors = bind_rows(olink_df_ligands_receptors, olink_df_ligands_receptors_NAs)
olink_df_ligands_receptors %>% head()
```

## Score sender/ligand-receiver/receptor interactions based on the alternative data modality

Important: the provided logFC values in the OLINK data contrast MIS-C versus healthy siblings (M-vs-S). We will stil need to add the reverse logFC for the contrast S-vs-M. 

```{r}
olink_df_ligand = prioritization_tables$group_prioritization_tbl %>% distinct(ligand) %>% left_join(olink_df_ligands_receptors %>% rename(ligand = gene)) %>% mutate(contrast = "M-S")
olink_df_reverse_ligand = olink_df_ligand %>% mutate(contrast = "S-M", logFC = -1*logFC)
olink_df_ligand = olink_df_ligand %>% bind_rows(olink_df_reverse_ligand)
olink_df_ligand %>% filter(ligand == "IFNG")
```

We will now score ligand-contrast combinations based on their OLINK logFC and p-values. We will do this in the exact same way as we did for differential expression at the RNA level.  

```{r}
olink_df_ligand = olink_df_ligand %>% mutate(pval_adapted = -log10(pval)*sign(logFC)) 
olink_df_ligand_scaled = olink_df_ligand %>% distinct(contrast, ligand, logFC, pval_adapted) %>% dplyr::mutate(scaled_lfc_ligand_OLINK = rank(logFC, ties.method = "average", na.last = FALSE)/max(rank(logFC, ties.method = "average", na.last = FALSE)), scaled_p_val_ligand_adapted_OLINK = rank(pval_adapted, ties.method = "average", na.last = FALSE)/max(rank(pval_adapted, ties.method = "average", na.last = FALSE))) 
olink_df_ligand_scaled = olink_df_ligand_scaled %>% distinct(contrast, ligand, scaled_lfc_ligand_OLINK, scaled_p_val_ligand_adapted_OLINK) %>% dplyr::arrange(-scaled_p_val_ligand_adapted_OLINK)
olink_df_ligand_scaled
```



Now we will calculate the aggregated prioritization score with inclusion of the scaled logFC and scaled p-value-based score from the OLINK data. 
To do this, we first add these new scores to the existing prioritization_table and define the table of prioritization criteria and add the new criteria and their weights. It is recommended to give the new criteria the same weights as the original default. If you ran the analysis with the default weights (`scenario = "regular"` then these weights are 1)  

```{r}
prioritization_tables$group_prioritization_tbl = prioritization_tables$group_prioritization_tbl %>% inner_join(olink_df_ligand_scaled)
```

```{r}
regular_criteria_tbl = tibble(criterion = c("scaled_lfc_ligand","scaled_p_val_ligand_adapted","scaled_lfc_receptor","scaled_p_val_receptor_adapted", "max_scaled_activity", "scaled_pb_ligand", "scaled_pb_receptor", "fraction_expressing_ligand_receptor"), weight = NA, regularization_factor = c(0.5, 0.5, 0.5, 0.5, 1, 1, 1, 1)) # do not change this  - logFC and pval of ligand and receptor get 0.5 as regularization such that DE of the ligandcounts as one score, which is the sum of the scaled_score of the logFC and the p-value-based score. This also applies to the receptor.

new_criteria_tbl = tibble(criterion = c("scaled_lfc_ligand_OLINK","scaled_p_val_ligand_adapted_OLINK"), weight = c(1,1), regularization_factor = c(0.5, 0.5)) # 0.5 as regularization such that DE of the ligand at the protein level counts as one score, which is the sum of the scaled_score of the logFC and the p-value-based score. 
```

```{r}
prioritization_tables = add_extra_criterion(prioritization_tables, new_criteria_tbl, regular_criteria_tbl, scenario = "regular") 
```

```{r}
prioritization_tables$group_prioritization_tbl %>% head(20)
```

### Add additional ligand activity criterion

As last point, we will demonstrate how we can also use the serum proteomics data to determine ligand signatures in the entire serum proteome. 

Calculate ligand activities on the serum proteome
```{r}
logFC_threshold = 1
pval_threshold = 0.05
top_n_target = 250
verbose = TRUE
olink_df = olink_df %>% mutate(contrast = contrast_tbl$contrast %>% .[1])
  olink_df_reverse = olink_df %>% mutate(contrast = contrast_tbl$contrast %>% .[2], logFC = -1*logFC)
  de_output_tidy = olink_df %>% bind_rows(olink_df_reverse)

  background_expressed_genes = de_output_tidy$gene %>% unique() %>% dplyr::intersect(rownames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[rownames(ligand_target_matrix) %in% background_expressed_genes, ]
  ligands = colnames(ligand_target_matrix)
      
  geneset_vs_ligand_activities = list()
  ligand_activities_targets_geneset = list()
  for(i in seq(length(de_output_tidy$contrast %>% unique()))){
    contrast_oi = de_output_tidy$contrast %>% unique() %>% .[i]
    de_tbl_geneset = de_output_tidy %>% dplyr::filter(contrast == contrast_oi) %>% dplyr::filter(logFC >= logFC_threshold & pval <= pval_threshold)
    geneset_oi = de_tbl_geneset %>% dplyr::pull(gene) %>% unique() %>% dplyr::intersect(rownames(ligand_target_matrix))
    
    if(verbose == TRUE){
      print("contrast_oi:")
      print(contrast_oi)
    }
    if(length(geneset_oi) > 0){
      if(verbose == TRUE){
        print("Number of upregulated DE genes (gene set of interest): ")
        print(length(geneset_oi))
      }    

      ligand_activities = nichenetr::predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = ligands)

      ligand_activities = ligand_activities %>% dplyr::mutate(contrast = contrast_oi) %>% tidyr::drop_na() %>% dplyr::rename(ligand = test_ligand, activity = aupr_corrected) %>% dplyr::select(-pearson, -auroc, -aupr)
      ligand_target_df = ligand_activities$ligand %>% unique() %>% lapply(nichenetr::get_weighted_ligand_target_links, geneset_oi, ligand_target_matrix, top_n_target) %>% dplyr::bind_rows() %>% dplyr::mutate(contrast = contrast_oi) %>% dplyr::rename(ligand_target_weight = weight)
      ligand_activities = ligand_activities %>% dplyr::inner_join(ligand_target_df)  %>% dplyr::mutate(direction_regulation = "up") %>% dplyr::group_by(contrast) %>% dplyr::mutate(activity_scaled = nichenetr::scaling_zscore(activity), direction_regulation = "up")
    } else {
      warning(paste0("In condition ", contrast_oi, " there seem to be no upregulated DE genes - so ligand activities will be NA. Please check the DE output."))
      ligand_activities = tibble(ligand = ligands, activity = NA, contrast = contrast_oi, target = NA, direction_regulation = "up",  ligand_target_weight = NA, activity_scaled = NA)
    }

    de_genes_df = de_tbl_geneset %>% dplyr::mutate(contrast = contrast_oi) 
    
    ligand_activities_targets_geneset[[i]] = list(ligand_activities = ligand_activities, de_genes_df = de_genes_df)
  }
  
  ligand_activities = ligand_activities_targets_geneset %>% purrr::map("ligand_activities") %>% dplyr::bind_rows()
  de_genes_df = ligand_activities_targets_geneset %>% purrr::map("de_genes_df") %>% dplyr::bind_rows()
```

```{r}
# OLINK lignand activity prioritization
OLINK_ligand_activity_prioritization_up = ligand_activities %>% dplyr::ungroup() %>% dplyr::distinct(contrast, ligand, activity_scaled) %>% dplyr::mutate(scaled_activity_scaled_up_OLINK = nichenetr::scale_quantile_adapted(activity_scaled, outlier_cutoff = 0.01)) %>% dplyr::arrange(-scaled_activity_scaled_up_OLINK) %>% dplyr::distinct(contrast, ligand, scaled_activity_scaled_up_OLINK) 
```

```{r}
prioritization_tables$group_prioritization_tbl = prioritization_tables$group_prioritization_tbl %>% inner_join(OLINK_ligand_activity_prioritization_up)
```

```{r}
regular_criteria_tbl = bind_rows(regular_criteria_tbl, new_criteria_tbl)

new_criteria_tbl = tibble(criterion = c("scaled_activity_scaled_up_OLINK"), weight = 1, regularization_factor = 1) 
```

```{r}
prioritization_tables = add_extra_criterion(prioritization_tables, new_criteria_tbl, regular_criteria_tbl, scenario = "regular") 
```

```{r}
prioritization_tables$group_prioritization_tbl %>% head(20)
```

# Step 7: Add information on prior knowledge and expression correlation between LR and target expression. 

In multi-sample datasets, we have the opportunity to look whether expression of ligand-receptor across all samples is correlated with the expression of their by NicheNet predicted target genes. This is what we will do with the following line of code:

```{r}
lr_target_prior_cor = lr_target_prior_cor_inference(prioritization_tables$group_prioritization_tbl$receiver %>% unique(), abundance_expression_info, celltype_de, grouping_tbl, prioritization_tables, ligand_target_matrix, logFC_threshold = logFC_threshold, p_val_threshold = p_val_threshold, p_val_adj = p_val_adj)
```

# Step 8: Save all the output of MultiNicheNet 

To avoid needing to redo the analysis later.
All the output written down here is sufficient to make all in-built downstream visualizations.

```{r}
path = "./"

multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de =  sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  ) 
multinichenet_output = make_lite_output(multinichenet_output)

save = FALSE
if(save == TRUE){
  saveRDS(multinichenet_output, paste0(path, "multinichenet_output.rds"))

}
```


# Visualization of the results of the cell-cell communication analysis

In a first instance, we will look at the broad overview of prioritized interactions via condition-specific Circos plots.

## Circos plot of top-prioritized links

We will look here at the top 50 predictions across all contrasts, senders, and receivers of interest.

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, rank_per_group = FALSE)
```

```{r, fig.width=15, fig.height=12}
prioritized_tbl_oi = multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

## Visualization of scaled ligand-receptor pseudobulk products and ligand activity

Now we will visualize per sample the scaled product of ligand and receptor expression. Samples that were left out of the DE analysis are indicated with a smaller dot (this helps to indicate the samples that did not contribute to the calculation of the logFC, and thus not contributed to the final prioritization)

We will first visualize the interactions specific for the Sibling group that are part of the top50 differential interactions as shown in the above circos plots

```{r}
group_oi = "S"
```

```{r, fig.height=5, fig.width=13}
prioritized_tbl_oi_S_50 = prioritized_tbl_oi_all %>% filter(group == group_oi)

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_S_50)
plot_oi
```

We will now check the MIS-C specific interactions

```{r}
group_oi = "M"
```

```{r, fig.height=13, fig.width=15}
prioritized_tbl_oi_M_50 = prioritized_tbl_oi_all %>% filter(group == group_oi)

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_M_50)
plot_oi
```

Whereas these are the classical plots with default multinichenetr code, we can also adapt this function to visualize the additional criteria that we used for prioritization. 

We will show example code to make this multi-panel plot step-by-step

First: ligand-receptor pseudobulk product expression panel

```{r, fig.width=10, fig.height=8}
# ligand-receptor pseudobulk product expression panel
sample_data = prioritization_tables$sample_prioritization_tbl %>% dplyr::filter(id %in% prioritized_tbl_oi$id) %>% dplyr::mutate(sender_receiver = paste(sender, receiver, sep = " --> "), lr_interaction = paste(ligand, receptor, sep = " - "))   %>%  dplyr::arrange(receiver) %>% dplyr::group_by(receiver) %>%  dplyr::arrange(sender, .by_group = TRUE)
sample_data = sample_data %>% dplyr::mutate(sender_receiver = factor(sender_receiver, levels = sample_data$sender_receiver %>% unique()))
  
keep_sender_receiver_values = c(0.25, 0.9, 1.75, 4)
names(keep_sender_receiver_values) = levels(sample_data$keep_sender_receiver)
  
p1 = sample_data %>%
    ggplot(aes(sample, lr_interaction, color = scaled_LR_pb_prod, size = keep_sender_receiver)) +
    geom_point() +
    facet_grid(sender_receiver~group, scales = "free", space = "free", switch = "y") +
    scale_x_discrete(position = "top") +
    # xlab("Ligand-Receptor expression in samples\n\n") +
    theme_light() +
    theme(
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      # axis.title.x = element_text(face = "bold", size = 11),       axis.title.y = element_blank(),
      axis.text.y = element_text(face = "bold.italic", size = 9),
      axis.text.x = element_text(size = 9,  angle = 90,hjust = 0),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(0.40, "lines"),
      panel.spacing.y = unit(0.25, "lines"),
      strip.text.x.top = element_text(size = 10, color = "black", face = "bold", angle = 0),
      strip.text.y.left = element_text(size = 9, color = "black", face = "bold", angle = 0),
      strip.background = element_rect(color="darkgrey", fill="whitesmoke", size=1.5, linetype="solid")
    ) + labs(color = "Scaled L-R\npseudobulk exprs product", size= "Sufficient presence\nof sender & receiver") + 
    scale_size_manual(values = keep_sender_receiver_values)
  max_lfc = abs(sample_data$scaled_LR_pb_prod) %>% max()
  custom_scale_fill = scale_color_gradientn(colours = RColorBrewer::brewer.pal(n = 7, name = "RdBu") %>% rev(),values = c(0, 0.350, 0.4850, 0.5, 0.5150, 0.65, 1),  limits = c(-1*max_lfc, max_lfc))
  
  p1 = p1 + custom_scale_fill
p1
```  
Second: scaled ligand activity based on RNA data

```{r, fig.width=4, fig.height=8}
# scaled ligand activity based on RNA
group_data = prioritization_tables$group_prioritization_table_source  %>% 
    dplyr::mutate(sender_receiver = paste(sender, receiver, sep = " --> "), lr_interaction = paste(ligand, receptor, sep = " - "))  %>% 
    dplyr::distinct(id, sender, receiver, sender_receiver, ligand, receptor, lr_interaction, group, activity_scaled, direction_regulation, prioritization_score ) %>% dplyr::filter(id %in% sample_data$id) %>%  dplyr::arrange(receiver) %>% dplyr::group_by(receiver) %>%  dplyr::arrange(sender, .by_group = TRUE)

  group_data = group_data %>% dplyr::mutate(sender_receiver = factor(sender_receiver, levels = group_data$sender_receiver %>% unique()))

  p2 = group_data %>%
    # ggplot(aes(receiver, lr_interaction, color = activity_scaled, size = activity)) +
    # geom_point() +
    ggplot(aes(direction_regulation , lr_interaction, fill = activity_scaled)) +
    geom_tile(color = "whitesmoke") +
    facet_grid(sender_receiver~group, scales = "free", space = "free") +
    scale_x_discrete(position = "top") +
    # xlab("Ligand activities in receiver cell types\n\n") +
    theme_light() +
    theme(
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      # axis.title.x = element_text(face = "bold", size = 11),       axis.title.y = element_blank(),
      # axis.text.y = element_blank(),
      axis.text.y = element_text(face = "bold.italic", size = 9),
      axis.text.x = element_text(size = 9,  angle = 90,hjust = 0),
      strip.text.x.top = element_text(angle = 0),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(0.20, "lines"),
      panel.spacing.y = unit(0.25, "lines"),
      strip.text.x = element_text(size = 10, color = "black", face = "bold"),
      strip.text.y = element_blank(),
      strip.background = element_rect(color="darkgrey", fill="whitesmoke", size=1.5, linetype="solid")
    ) + labs(fill = "Scaled Ligand\nActivity RNA")
  max_activity = abs(group_data$activity_scaled) %>% max(na.rm = TRUE)
  custom_scale_fill = scale_fill_gradientn(colours = c("white", RColorBrewer::brewer.pal(n = 7, name = "PuRd") %>% .[-7]),values = c(0, 0.51, 0.575, 0.625, 0.675, 0.725, 1),  limits = c(-1*max_activity, max_activity))
  
  p2 = p2 + custom_scale_fill
p2
```    

Third: scaled ligand activity based on OLINK data

```{r, fig.width=3, fig.height=8}
OLINK_ligand_activities_processed = ligand_activities %>% distinct(ligand, direction_regulation, activity_scaled, contrast) %>% inner_join(contrast_tbl) %>% rename(OLINK_activity_scaled = activity_scaled) %>% mutate(direction_regulation = factor(direction_regulation, levels = c("up","down")))

group_data = group_data %>% left_join(OLINK_ligand_activities_processed)
  
p3 = group_data %>% filter(direction_regulation == "up") %>%
    # ggplot(aes(receiver, lr_interaction, color = activity_scaled, size = activity)) +
    # geom_point() +
    ggplot(aes(direction_regulation , lr_interaction, fill = OLINK_activity_scaled)) +
    geom_tile(color = "whitesmoke") +
    facet_grid(sender_receiver~group, scales = "free", space = "free") +
    scale_x_discrete(position = "top") +
    # xlab("Ligand activities in receiver cell types\n\n") +
    theme_light() +
    theme(
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      # axis.title.x = element_text(face = "bold", size = 11),       axis.title.y = element_blank(),
      # axis.text.y = element_blank(),
      axis.text.y = element_text(face = "bold.italic", size = 0),
      axis.text.x = element_text(size = 9,  angle = 90,hjust = 0),
      strip.text.x.top = element_text(angle = 0),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(0.20, "lines"),
      panel.spacing.y = unit(0.25, "lines"),
      strip.text.x = element_text(size = 10, color = "black", face = "bold"),
      strip.text.y = element_blank(),
      strip.background = element_rect(color="darkgrey", fill="whitesmoke", size=1.5, linetype="solid")
    ) + labs(fill = "Scaled Ligand\nActivity OLINK")
  max_activity = abs(group_data$OLINK_activity_scaled) %>% max(na.rm = TRUE)
  custom_scale_fill = scale_fill_gradientn(colours = c("white", RColorBrewer::brewer.pal(n = 7, name = "BuGn") %>% .[-7]),values = c(0, 0.51, 0.575, 0.625, 0.675, 0.725, 1),  limits = c(-1*max_activity, max_activity))
  
  p3 = p3 + custom_scale_fill
p3  
```      
  
Fourth: OLINK logFC MISC-vs-Sibling

```{r, fig.width=3, fig.height=8}
  
group_data_olink = inner_join(
    group_data %>% left_join(olink_df %>% filter(contrast == "M-S") %>% select(gene, olink_type, logFC) %>% dplyr::rename(ligand = gene, `ligand ` = logFC)) %>% left_join(olink_df %>% select(gene, olink_type, logFC) %>% dplyr::rename(receptor = gene, `receptor ` = logFC))  %>% gather(LR, logFC,`ligand `:`receptor `),
    group_data %>% left_join(olink_df %>% filter(contrast == "M-S") %>% select(gene, olink_type, pval) %>% dplyr::rename(ligand = gene, `ligand ` = pval)) %>% left_join(olink_df %>% select(gene, olink_type, pval) %>% dplyr::rename(receptor = gene, `receptor ` = pval))  %>% gather(LR, pval,`ligand `:`receptor `) %>% mutate(neg_log_pval = -log10(pval))
  )
  group_data_olink = group_data_olink %>% mutate(olink_type = "Diorio")
  
  p_olink = group_data_olink %>% 
    ggplot(aes(LR , lr_interaction, color = logFC, size = neg_log_pval)) +
    geom_point() +
    facet_grid(sender_receiver ~ olink_type, scales = "free", space = "free") +
    scale_x_discrete(position = "top") +
    theme_light() +
    theme(
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.text.x = element_text(size = 9,  angle = 90,hjust = 0),
      strip.text.x.top = element_text(angle = 0),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(0.20, "lines"),
      panel.spacing.y = unit(0.25, "lines"),
      strip.text.x = element_text(size = 10, color = "black", face = "bold"),
      strip.text.y = element_blank(),
      strip.background = element_rect(color="darkgrey", fill="whitesmoke", size=1.5, linetype="solid")
    ) + labs(color = "OLink LogFC M-vs-S") + labs(size = "OLink -log10(pval)")
  max_lfc = abs(group_data_olink$logFC) %>% max(na.rm = TRUE)
  custom_scale_fill = scale_color_gradientn(colours = RColorBrewer::brewer.pal(n = 7, name = "RdBu") %>% rev(),values = c(0, 0.30, 0.40, 0.5, 0.60, 0.70, 1),  limits = c(-1*max_lfc, max_lfc))
  
  p_olink = p_olink + custom_scale_fill + scale_size_binned_area(max_size = 4)
p_olink
```      
And now we will put everything together:
```{r, fig.width=15, fig.height=10}
p = patchwork::wrap_plots(
      p1,p2,p3,p_olink,
      nrow = 1,guides = "collect",
      widths = c(sample_data$sample %>% unique() %>% length(), 1*(sample_data$group %>% unique() %>% length()), 1*(sample_data$group %>% unique() %>% length()), 2*length(unique(olink_df$olink_type)))
)
p
```      

