---
title: "MultiNicheNet analysis: MIS-C threewise comparison - step-by-step"
author: "Robin Browaeys"
date: "2023-06-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MultiNicheNet analysis: MIS-C threewise comparison - step-by-step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- github markdown built using 
rmarkdown::render("vignettes/basic_analysis_steps_MISC.Rmd", output_format = "github_document")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

In this vignette, you can learn how to perform an all-vs-all MultiNicheNet analysis. In this vignette, we start from one SingleCellExperiment object containing cells from both sender and receiver cell types and from different patients.

A MultiNicheNet analysis can be performed if you have multi-sample, multi-group single-cell data. MultiNicheNet will look for cell-cell communication between the cell types in your data for each sample, and compare the cell-cell communication patterns between the groups of interest. Therefore, the absolute minimum of meta data you need to have, are following columns indicating for each cell: the **group**, **sample** and **cell type**.
 
As example expression data of interacting cells, we will here use scRNAseq data of immune cells in MIS-C patients and healthy siblings from this paper of Hoste et al.: [TIM3+ TRBV11-2 T cells and IFNγ signature in patrolling monocytes and CD16+ NK cells delineate MIS-C](https://rupress.org/jem/article/219/2/e20211381/212918/TIM3-TRBV11-2-T-cells-and-IFN-signature-in) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6362434.svg)](https://doi.org/10.5281/zenodo.6362434)
. MIS-C (multisystem inflammatory syndrome in children) is a novel rare immunodysregulation syndrome that can arise after SARS-CoV-2 infection in children. We will use NicheNet to explore immune cell crosstalk enriched in MIS-C compared to healthy siblings. 

In this vignette, we will prepare the data and analysis parameters, and then perform the MultiNicheNet analysis. 

The different steps of the MultiNicheNet analysis are the following:

* 0. Preparation of the analysis: load packages, NicheNet LR network & ligand-target matrix, single-cell expression data, and define main settings of the MultiNicheNet analysis

* 1. Extract cell type abundance and expression information from receiver and sender cell types, and link this expression information for ligands of the sender cell types to the corresponding receptors of the receiver cell types 

* 2. Perform genome-wide differential expression analysis of receiver and sender cell types to define DE genes between the conditions of interest. Based on this analysis, we can define the logFC/p-value of ligands in senders and receptors in receivers, and define the set of affected target genes in the receiver.

* 3. Predict NicheNet ligand activities and NicheNet ligand-target links based on these differential expression results

* 4. Use the information collected above to prioritize all sender-ligand---receiver-receptor pairs.

* 5. Calculate correlation in expression between ligand-receptor pairs and their predicted target genes

In this vignette, we will demonstrate all these steps one by one.

After the MultiNicheNet analysis is done, we will explore the output of the analysis with different ways of visualization. 

# Step 0: Preparation of the analysis: load packages, NicheNet LR network & ligand-target matrix, single-cell expression data

## Step 0.1: Load required packages and NicheNet ligand-receptor network and ligand-target matrix

```{r}
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(nichenetr)
library(multinichenetr)
```

The Nichenet v2 networks and matrices for both mouse and human can be downloaded from Zenodo [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.7074291.svg)](https://doi.org/10.5281/zenodo.7074291). 

We will read these object in for human because our expression data is of human patients. 
Gene names are here made syntactically valid via `make.names()` to avoid the loss of genes (eg H2-M3) in downstream visualizations.

```{r}
organism = "human"
if(organism == "human"){
  lr_network_all = readRDS(url("https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds")) %>% mutate(ligand = convert_alias_to_symbols(ligand, organism = organism), receptor = convert_alias_to_symbols(receptor, organism = organism))
  lr_network = lr_network_all  %>% mutate(ligand = make.names(ligand), receptor = make.names(receptor)) %>% distinct(ligand, receptor)
  ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% convert_alias_to_symbols(organism = organism) %>% make.names()
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
} else if(organism == "mouse"){
  lr_network_all = readRDS(url("https://zenodo.org/record/10229222/files/lr_network_mouse_allInfo_30112033.rds")) %>% mutate(ligand = convert_alias_to_symbols(ligand, organism = organism), receptor = convert_alias_to_symbols(receptor, organism = organism))
  lr_network = lr_network_all  %>% mutate(ligand = make.names(ligand), receptor = make.names(receptor)) %>% distinct(ligand, receptor)
  ligand_target_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% convert_alias_to_symbols(organism = organism) %>% make.names()
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
}
```

## Step 0.2: Read in SingleCellExperiment Objects 

In this vignette, sender and receiver cell types are in the same SingleCellExperiment object, which we will load here. In this vignette, we will load in a subset of the scRNAseq data of the MIS-C [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.8010790.svg)](https://doi.org/10.5281/zenodo.8010790). For the sake of demonstration, this subset only contains 3 cell types. These celltypes are some of the cell types that were found to be most interesting related to MIS-C according to Hoste et al. 

If you start from a Seurat object, you can convert it easily to a SingleCellExperiment via `sce = Seurat::as.SingleCellExperiment(seurat_obj, assay = "RNA")`.

Because the NicheNet 2.0. networks are in the most recent version of the official gene symbols, we will make sure that the gene symbols used in the expression data are also updated (= converted from their "aliases" to official gene symbols). Afterwards, we will make them again syntactically valid. 

```{r}
sce = readRDS(url("https://zenodo.org/record/8010790/files/sce_subset_misc.rds"))
sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()
```

## Step 0.3: Prepare settings of the MultiNicheNet cell-cell communication analysis

### Define in which metadata columns we can find the **group**, **sample** and **cell type** IDs

In this case study, we want to study differences in cell-cell communication patterns between MIS-C patients (M), their healthy siblings (S) and adult patients with severe covid (A). The meta data columns that indicate this disease status is `MIS.C.AgeTier`. 

Cell type annotations are indicated in the `Annotation_v2.0` column, and the sample is indicated by the `ShortID` column. 
If your cells are annotated in multiple hierarchical levels, we recommend using a high level in the hierarchy. This for 2 reasons: 1) MultiNicheNet focuses on differential expression and not differential abundance, and 2) there should be sufficient cells per sample-celltype combination.

If you would have batch effects or covariates you can correct for, you can define this here as well. 

Important: for categorical covariates and batches, there should be at least one sample for every group-batch combination. If one of your groups/conditions lacks a certain level of your batch, you won't be able to correct for the batch effect because the model is then not able to distinguish batch from group/condition effects.

Important: the column names of group, sample, cell type, batches and covariates should be syntactically valid (`make.names`)

Important: All group, sample, cell type, batch and covariate names should be syntactically valid as well (`make.names`) (eg through `SummarizedExperiment::colData(sce)$ShortID = SummarizedExperiment::colData(sce)$ShortID %>% make.names()`)

```{r}
sample_id = "ShortID"
group_id = "MIS.C.AgeTier"
celltype_id = "Annotation_v2.0"
covariates = NA
batches = NA
```

Sender and receiver cell types also need to be defined. Both are here all cell types in the dataset because we are interested in an All-vs-All analysis.

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
```

If the user wants it, it is possible to use only a subset of senders and receivers. Senders and receivers can be entirely different, but also overlapping, or the same. If you don't use all the cell types in your data, we recommend to continue with a subset of your data.

```{r}
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% c(senders_oi, receivers_oi)]
```

Now we will go to the first real step of the MultiNicheNet analysis

# Step 1: Extract cell type abundance and expression information from receiver and sender cell types, and link this expression information for ligands of the sender cell types to the corresponding receptors of the receiver cell types

Since MultiNicheNet will infer group differences at the sample level for each cell type (currently via Muscat - pseudobulking + EdgeR), we need to have sufficient cells per sample of a cell type, and this for both groups. In the following analysis we will set this minimum number of cells per cell type per sample at 10 (recommended minimum). 

```{r}
min_cells = 10
```

Now we will calculate abundance and expression information for each cell type / sample / group combination with the following functions. In the output of this function, you can also find some 'Cell type abundance diagnostic plots' that will the users which celltype-sample combinations will be left out later on for DE calculation because the nr of cells is lower than de defined minimum defined here above. If too many celltype-sample combinations don't pass this threshold, we recommend to define your cell types in a more general way (use one level higher of the cell type ontology hierarchy) (eg TH17 CD4T cells --> CD4T cells).

```{r}
abundance_info = get_abundance_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, min_cells = min_cells, senders_oi = senders_oi, receivers_oi = receivers_oi, batches = batches)
```

First, we will check the cell type abundance diagnostic plots.

### Interpretation of cell type abundance information

The first plot visualizes the number of cells per celltype-sample combination, and indicates which combinations are removed during the DE analysis because there are less than `min_cells` in the celltype-sample combination. 

```{r, fig.width=15, fig.height=12}
abundance_info$abund_plot_sample
```
The red dotted line indicates the required minimum of cells as defined above in `min_cells`. We can see here that some sample-celltype combinations are left out. For the DE analysis in the next step, only cell types will be considered if there are at least two samples per group with a sufficient number of cells.

__Important__: Based on the cell type abundance diagnostics, we recommend users to change their analysis settings if required (eg changing cell type annotation level, batches, ...), before proceeding with the rest of the analysis.

as we can see here: no condition-specific cell types. In case you would have them...

# Step 2: Perform genome-wide differential expression analysis of receiver and sender cell types to define DE genes between the conditions of interest. Based on this analysis, we can define the logFC/p-value of ligands in senders and receptors in receivers, and define the set of affected target genes in the receiver.

Now we will go over to the multi-group, multi-sample differential expression (DE) analysis (also called 'differential state' analysis by the developers of Muscat).

### Define the contrasts and covariates of interest for the DE analysis.

Here, we want to compare each patient group to the other groups, so the MIS-C (M) group vs healthy control siblins (S) and adult COVID19 patients (A) etcetera. 
To do this comparison, we need to set the following contrasts:

```{r}
contrasts_oi = c("'M-(S+A)/2','S-(M+A)/2','A-(S+M)/2'")
```

__Very Important__ Note the format to indicate the contrasts! This formatting should be adhered to very strictly, and white spaces are not allowed!  Check `?get_DE_info` for explanation about how to define this well. The most important things are that: each contrast is surrounded by single quotation marks, contrasts are separated by a comma without any whitespace, and alle contrasts together are surrounded by double quotation marks. If you compare against two groups, you should divide by 2, if you compare against three groups, you should divide by 3 etcetera.

For downstream visualizations and linking contrasts to their main group, you need to run the following:

```{r}
contrast_tbl = tibble(contrast = 
                        c("M-(S+A)/2","S-(M+A)/2", "A-(S+M)/2"), 
                      group = c("M","S","A"))
```

If you want to compare only two groups (eg M vs S), you can do like this:
`contrasts_oi = c("'M-S','S-M'") `
`contrast_tbl = tibble(contrast = c("M-S","S-M"), group = c("M","S"))`

### Perform the DE analysis for each cell type.

First define expressed genes
```{r}
fraction_cutoff = 0.05
min_sample_prop = 0.50
frq_list = get_frac_exprs(sce = sce, sample_id = sample_id, celltype_id =  celltype_id, group_id = group_id, batches = batches, min_cells = min_cells, fraction_cutoff = fraction_cutoff, min_sample_prop = min_sample_prop)
```

```{r}
DE_info = get_DE_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, batches = batches, covariates = covariates, contrasts_oi = contrasts_oi, min_cells = min_cells, expressed_df = frq_list$expressed_df)
```

### Check DE results

Table with logFC and p-values for each gene-celltype-contrast:

```{r}
DE_info$celltype_de$de_output_tidy %>% arrange(p_adj) %>% head()
```

```{r}
celltype_de = DE_info$celltype_de$de_output_tidy
```

In the next step, we will combine the DE information of senders and receivers by linking their ligands and receptors together:

### Combine DE information for ligand-senders and receptors-receivers (similar to step1 - `abundance_expression_info$sender_receiver_info`)

```{r}
sender_receiver_de = combine_sender_receiver_de(
  sender_de = celltype_de,
  receiver_de = celltype_de,
  senders_oi = senders_oi,
  receivers_oi = receivers_oi,
  lr_network = lr_network
)
```

```{r}
sender_receiver_de %>% head(20)
```

# Calculate (pseudobulk) expression averages

```{r}
abundance_expression_info = process_abundance_expression_info(sce = sce, sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, min_cells = min_cells, senders_oi = senders_oi, receivers_oi = receivers_oi, lr_network = lr_network, batches = batches, frq_list = frq_list, abundance_info = abundance_info)
```

# Step 3: Predict NicheNet ligand activities and NicheNet ligand-target links based on these differential expression results

## Define the parameters for the NicheNet ligand activity analysis 

Here, we need to define the thresholds that will be used to consider genes as differentially expressed or not (logFC, p-value, decision whether to use adjusted or normal p-value, minimum fraction of cells that should express a gene in at least one sample in a group, whether to use the normal p-values or empirical p-values). 

NicheNet ligand activity will then be calculated as the enrichment of predicted target genes of ligands in this set of DE genes compared to the genomic background. Here we choose for a minimum logFC of 0.50, maximum p-value of 0.05, and minimum fraction of expression of 0.05. 

```{r}
logFC_threshold = 0.50
p_val_threshold = 0.05
```

We will here choose for applying the p-value cutoff on the normal p-values, and not on the p-values corrected for multiple testing. This choice was made here because this dataset has only a few samples per group and we might have a lack of statistical power due to pseudobulking. In case of more samples per group, and a sufficient high number of DE genes per group-celltype (> 50), we would recommend using the adjusted p-values.

```{r}
# p_val_adj = TRUE 
p_val_adj = FALSE 
```

For the NicheNet ligand-target inference, we also need to select which top n of the predicted target genes will be considered (here: top 250 targets per ligand).

```{r}
top_n_target = 250
```

The NicheNet ligand activity analysis can be run in parallel for each receiver cell type, by changing the number of cores as defined here. This is only recommended if you have many receiver cell type. 

```{r}
verbose = TRUE
cores_system = 8
n.cores = min(cores_system, sender_receiver_de$receiver %>% unique() %>% length()) # use one core per receiver cell type
```

## Run the NicheNet ligand activity analysis 

(this might take some time)
```{r}
ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(get_ligand_activities_targets_DEgenes(
  receiver_de = celltype_de,
  receivers_oi = receivers_oi,
  ligand_target_matrix = ligand_target_matrix,
  logFC_threshold = logFC_threshold,
  p_val_threshold = p_val_threshold,
  p_val_adj = p_val_adj,
  top_n_target = top_n_target,
  verbose = verbose, 
  n.cores = n.cores
)))
```

Check the DE genes used for the activity analysis

```{r}
ligand_activities_targets_DEgenes$de_genes_df %>% head(20)
```

Check the output of the activity analysis

```{r}
ligand_activities_targets_DEgenes$ligand_activities %>% head(20)
```

# Step 4: Use the information collected above to prioritize all sender-ligand---receiver-receptor pairs.

In the 3 previous steps, we calculated expression, differential expression and NicheNet activity information. Now we will combine these different types of information in one prioritization scheme.

MultiNicheNet allows the user to define the weights of the following criteria to prioritize ligand-receptor interactions:

* Upregulation of the ligand in a sender cell type and/or upregulation of the receptor in a receiver cell type - in the condition of interest. : `de_ligand` and `de_receptor`
* Sufficiently high expression levels of ligand and receptor in many samples of the same group (to mitigate the influence of outlier samples). : `frac_exprs_ligand_receptor`
* Cell-type and condition specific expression of the ligand in the sender cell type and receptor in the receiver cell type (to mitigate the influence of upregulated but still relatively weakly expressed ligands/receptors) : `exprs_ligand` and `exprs_receptor`
* High NicheNet ligand activity, to further prioritize ligand-receptor pairs based on their predicted effect of the ligand-receptor interaction on the gene expression in the receiver cell type : `activity_scaled` 
* High relative abundance of sender and/or receiver in the condition of interest: `abund_sender` and `abund_receiver` (experimental feature - not recommended to give non-zero weights for default analyses)

The different properties of the sender-ligand---receiver-receptor pairs can be weighted according to the user's preference and insight in the dataset at hand. 

## Prepare grouping objects

Make necessary grouping data frame

```{r}
sender_receiver_tbl = sender_receiver_de %>% dplyr::distinct(sender, receiver)

metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

if(!is.na(batches)){
  grouping_tbl = metadata_combined[,c(sample_id, group_id, batches)] %>% tibble::as_tibble() %>% dplyr::distinct()
  colnames(grouping_tbl) = c("sample","group",batches)
} else {
  grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% tibble::as_tibble() %>% dplyr::distinct()
  colnames(grouping_tbl) = c("sample","group")
}

```

Crucial note: grouping_tbl: group should be the same as in the contrast_tbl, and as in the expression info tables! Rename accordingly if this would not be the case. If you followed the guidelines of this tutorial closely, there should be no problem.

## Run the prioritization 

```{r}
  prioritization_tables = suppressMessages(generate_prioritization_tables(
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    contrast_tbl = contrast_tbl,
    sender_receiver_tbl = sender_receiver_tbl,
    grouping_tbl = grouping_tbl,
    scenario = "regular", # all prioritization criteria will be weighted equally
    fraction_cutoff = fraction_cutoff, 
    abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
    abundance_data_sender = abundance_expression_info$abundance_data_sender,
    ligand_activity_down = FALSE # use only upregulatory ligand activities to prioritize -- if you want to consider max of up and down: set this to TRUE
  ))
```

Check the output tables

First: group-based summary table

```{r}
prioritization_tables$group_prioritization_tbl %>% head(20)
```


# Step 5: Add information on prior knowledge and expression correlation between LR and target expression. 

In multi-sample datasets, we have the opportunity to look whether expression of ligand-receptor across all samples is correlated with the expression of their by NicheNet predicted target genes. This is what we will do with the following line of code:

```{r}
lr_target_prior_cor = lr_target_prior_cor_inference(prioritization_tables$group_prioritization_tbl$receiver %>% unique(), abundance_expression_info, celltype_de, grouping_tbl, prioritization_tables, ligand_target_matrix, logFC_threshold = logFC_threshold, p_val_threshold = p_val_threshold, p_val_adj = p_val_adj)
```

# Save all the output of MultiNicheNet 

To avoid needing to redo the analysis later.
All the output written down here is sufficient to make all in-built downstream visualizations.

```{r}
path = "./"

multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de =  sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  ) 
multinichenet_output = make_lite_output(multinichenet_output)

save = FALSE
if(save == TRUE){
  saveRDS(multinichenet_output, paste0(path, "multinichenet_output.rds"))

}
```


# Visualization of the results of the cell-cell communication analysis

In a first instance, we will look at the broad overview of prioritized interactions via condition-specific Circos plots.

## Circos plot of top-prioritized links

We will look here at the top 50 predictions across all contrasts, senders, and receivers of interest.

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, rank_per_group = FALSE)
```


```{r, fig.width=15, fig.height=12}
prioritized_tbl_oi = multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

## Visualization of scaled ligand-receptor pseudobulk products and ligand activity

Now we will visualize per sample the scaled product of ligand and receptor expression. Samples that were left out of the DE analysis are indicated with a smaller dot (this helps to indicate the samples that did not contribute to the calculation of the logFC, and thus not contributed to the final prioritization)

We will now check the top 50 interactions specific for the MIS-C group

```{r}
group_oi = "M"
```

```{r, fig.height=13, fig.width=15}
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, groups_oi = group_oi)
```

```{r, fig.height=13, fig.width=18}
plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_M_50)
plot_oi
```

Typically, there are way more than 50 differentially expressed and active ligand-receptor pairs per group across all sender-receiver combinations. Therefore it might be useful to zoom in on specific cell types as senders/receivers:

Eg M_Monocyte_CD16 as receiver:

```{r, fig.height=13, fig.width=15}
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, groups_oi = group_oi, receivers_oi = "M_Monocyte_CD16")

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_M_50)
plot_oi
```

Eg M_Monocyte_CD16 as sender:

```{r, fig.height=13, fig.width=15}
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, groups_oi = group_oi, senders_oi = "M_Monocyte_CD16")

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_M_50)
plot_oi
```

You can make these plots also for the other groups, like we will illustrate now for the S group

```{r}
group_oi = "S"
```

```{r, fig.height=13, fig.width=15}
prioritized_tbl_oi_S_50 = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 50, groups_oi = group_oi)

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_S_50)
plot_oi
```

__Note__: Use `make_sample_lr_prod_activity_batch_plots` if you have batches and want to visualize them on this plot!

## Visualization of ligand-activity, ligand-target links, and target gene expression

In another type of plot, we can visualize the ligand activities for a group-receiver combination, and show the predicted ligand-target links, and also the expression of the predicted target genes across samples.

For this, we now need to define a receiver cell type of interest. As example, we will take `L_T_TIM3._CD38._HLADR.` cells as receiver, and look at the top 20 senderLigand-receiverReceptor pairs with these cells as receiver.

```{r}
group_oi = "M"
receiver_oi = "L_T_TIM3._CD38._HLADR."
prioritized_tbl_oi_M_10 = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 10, groups_oi = group_oi, receivers_oi = receiver_oi)
```

```{r, fig.width=20, fig.height=10}
combined_plot = make_ligand_activity_target_plot(group_oi, receiver_oi, prioritized_tbl_oi_M_10, multinichenet_output$prioritization_tables, multinichenet_output$ligand_activities_targets_DEgenes, contrast_tbl, multinichenet_output$grouping_tbl, multinichenet_output$celltype_info, ligand_target_matrix, plot_legend = FALSE)
combined_plot
```

__Note__ Use `make_DEgene_dotplot_pseudobulk_batch` if you want to indicate the batch of each sample to the plot

What if there is a specific ligand you are interested in?

```{r}
group_oi = "M"
receiver_oi = "L_T_TIM3._CD38._HLADR."
ligands_oi = c("IFNG","IL15")
prioritized_tbl_ligands_oi = get_top_n_lr_pairs(multinichenet_output$prioritization_tables, 10000, groups_oi = group_oi, receivers_oi = receiver_oi) %>% filter(ligand %in% ligands_oi) # ligands should still be in the output tables of course
```

```{r, fig.width=20, fig.height=10}
combined_plot = make_ligand_activity_target_plot(group_oi, receiver_oi, prioritized_tbl_ligands_oi, multinichenet_output$prioritization_tables, multinichenet_output$ligand_activities_targets_DEgenes, contrast_tbl, multinichenet_output$grouping_tbl, multinichenet_output$celltype_info, ligand_target_matrix, plot_legend = FALSE)
combined_plot
```

## Visualization of expression-correlated target genes of ligand-receptor pairs

Before, we had calculated the correlation in expression between ligand-receptor pairs and DE genes. Now we will filter out correlated ligand-receptor --> target links that both show high expression correlation (spearman or activity > 0.50 in this example) and have some prior knowledge to support their link.

```{r}
group_oi = "M"
receiver_oi = "M_Monocyte_CD16"
lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>% inner_join(multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% distinct(ligand, target, direction_regulation, contrast)) %>% inner_join(contrast_tbl) %>% filter(group == group_oi, receiver == receiver_oi)
lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% filter(direction_regulation == "up") %>% filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% filter(direction_regulation == "down") %>% filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50)) # downregulation -- negative correlation
lr_target_prior_cor_filtered = bind_rows(lr_target_prior_cor_filtered_up, lr_target_prior_cor_filtered_down)
```

Now we will visualize the top correlated target genes for the LR pairs that are also in the top 50 LR pairs discriminating the groups from each other:

```{r}
prioritized_tbl_oi = get_top_n_lr_pairs(prioritization_tables, 50, groups_oi = group_oi, receivers_oi = receiver_oi)
```


```{r, fig.width=26, fig.height=16}
lr_target_correlation_plot = make_lr_target_correlation_plot(multinichenet_output$prioritization_tables, prioritized_tbl_oi,  lr_target_prior_cor_filtered , multinichenet_output$grouping_tbl, multinichenet_output$celltype_info, receiver_oi,plot_legend = FALSE)
lr_target_correlation_plot$combined_plot
```

You can also visualize the expression correlation in the following way for a selected LR pair and their targets:

```{r, fig.width=21, fig.height=6}
ligand_oi = "IFNG"
receptor_oi = "IFNGR2"
sender_oi = "L_T_TIM3._CD38._HLADR."
receiver_oi = "M_Monocyte_CD16"
lr_target_scatter_plot = make_lr_target_scatter_plot(multinichenet_output$prioritization_tables, ligand_oi, receptor_oi, sender_oi, receiver_oi, multinichenet_output$celltype_info, multinichenet_output$grouping_tbl, lr_target_prior_cor_filtered)
lr_target_scatter_plot
```

## Intercellular regulatory network systems view

In the plots before, we demonstrated that some DE genes have both expression correlation and prior knowledge support to be downstream of ligand-receptor pairs. Interestingly, some target genes can be ligands or receptors themselves. This illustrates that cells can send signals to other cells, who as a response to these signals produce signals themselves to feedback to the original sender cells, or who will effect other cell types. 

As last plot, we can generate a 'systems' view of these intercellular feedback and cascade processes than can be occuring between the different cell populations involved. In this plot, we will draw links between ligands of sender cell types their ligand/receptor-annotated target genes in receiver cell types. So links are ligand-target links (= gene regulatory links) and not ligand-receptor protein-protein interactions!

```{r}
prioritized_tbl_oi = get_top_n_lr_pairs(prioritization_tables, 250, rank_per_group = FALSE)

lr_target_prior_cor_filtered = prioritization_tables$group_prioritization_tbl$group %>% unique() %>% lapply(function(group_oi){
  lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>% inner_join(multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% distinct(ligand, target, direction_regulation, contrast)) %>% inner_join(contrast_tbl) %>% filter(group == group_oi)
  lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% filter(direction_regulation == "up") %>% filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
  lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% filter(direction_regulation == "down") %>% filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50))
  lr_target_prior_cor_filtered = bind_rows(lr_target_prior_cor_filtered_up, lr_target_prior_cor_filtered_down)
}) %>% bind_rows()

lr_target_df = lr_target_prior_cor_filtered  %>% distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation) 
```

```{r}
network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi)
network$links %>% head()
network$nodes %>% head()
```

```{r, fig.width=30, fig.height=12}
colors_sender["L_T_TIM3._CD38._HLADR."] = "pink" # the  original yellow background with white font is not very readable
network_graph = visualize_network(network, colors_sender)
network_graph$plot
```

An interesting follow-up question would be to find nodes that are at the top of the hierarchy in this network, and could thus explain the cascade of intercellular communication events we observe.

Using PageRank on the opposite-directionality graph can help identify important nodes that act as sources in the hierarchical structure. By running PageRank on the reversed graph (where edges point in the opposite direction), you can potentially pinpoint nodes that have significant influence or act as sources of information in your network.

We will exemplify for the M-group

```{r,}
library(igraph)

# Create the normal graph (normal directionality)
normal_edges <- get.data.frame(network_graph$network_igraph) %>% filter(group == "M") %>% .[, c("from", "to")]  
normal_graph <- graph_from_data_frame(normal_edges, directed = TRUE)
# Calculate PageRank on the normal graph
pagerank_scores <- page_rank(normal_graph)$vector %>% sort(decreasing = TRUE)
pr_score_tbl = tibble(node = names(pagerank_scores), score_normal = pagerank_scores) %>% arrange(-score_normal)

# Create the reversed graph (opposite directionality)
reversed_edges <- get.data.frame(network_graph$network_igraph) %>% filter(group == "M") %>% .[, c("to", "from")]  # Swap 'from' and 'to' columns
reversed_graph <- graph_from_data_frame(reversed_edges, directed = TRUE)
# Calculate PageRank on the reversed graph
reversed_pagerank_scores <- page_rank(reversed_graph)$vector %>% sort(decreasing = TRUE)

pr_score_tbl = pr_score_tbl %>% inner_join(
  tibble(node = names(reversed_pagerank_scores), score_reverse = reversed_pagerank_scores) 
) %>% arrange(-score_reverse)
pr_score_tbl
```

Let's zoom now in on the M-group network to check this

```{r}
network = infer_intercellular_regulatory_network(lr_target_df %>% filter(group == "M"), prioritized_tbl_oi %>% filter(group == "M"))
network$links %>% head()
network$nodes %>% head()
```

```{r, fig.width=30, fig.height=20}
colors_sender["L_T_TIM3._CD38._HLADR."] = "pink" # the  original yellow background with white font is not very readable
network_graph = visualize_network(network, colors_sender)
network_graph$plot
```

Interestingly, we can also use this network to further prioritize differential CCC interactions. Here we will assume that the most important LR interactions are the ones that are involved in this intercellular regulatory network. We can get these interactions as follows:
```{r}
network$prioritized_lr_interactions
```

```{r, fig.width=30, fig.height=12}
prioritized_tbl_oi_network = prioritized_tbl_oi %>% inner_join(network$prioritized_lr_interactions)
prioritized_tbl_oi_network
```

Visualize now the expression and activity of these interactions for the M group
```{r}
group_oi = "M"
```

```{r, fig.height=16, fig.width=18}
prioritized_tbl_oi_M = prioritized_tbl_oi_network %>% filter(group == group_oi)

plot_oi = make_sample_lr_prod_activity_plots(multinichenet_output$prioritization_tables, prioritized_tbl_oi_M)
plot_oi
```

## Investingating the data sources behind the ligand-receptor interactions

Now link the prioritized interactions to all the information of the ligand-receptor network. This can help to check which are the specific data sources behind the inferred interactions, which can aid in prioritizing interactions for further validation.  

Check OmnipathR for more information with respect to how to interpret the different metrics specific for the interactions coming from the Omnipath database.
```{r}
prioritized_tbl_oi_M %>% inner_join(lr_network_all) %>% arrange(-n_resources) # sort interactions based on nr of resources that document them
```
