---
title: "Multi-Sample Multi-condition Cell-Cell Communication Analysis via NicheNet: HNSCC application; Separate"
author: "Robin Browaeys"
date: "2021-04-09"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Sample Multi-condition Cell-Cell Communication Analysis via NicheNet: HNSCC application; Separate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- github markdown built using 
rmarkdown::render("vignettes/basic_analysis_separate.Rmd", output_format = "github_document")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

In this vignette, you can learn how to perform an all-vs-all MultiNicheNet analysis. In this vignette, we start from two Seurat objects: one seurat object containing the sender cell types, and one seurat object with receiver cell types. (for demonstration purposes, we start from the same object, that we then split based on sender/receiver cell type)

As example expression data of interacting cells, we will use data from Puram et al. to explore intercellular communication in the tumor microenvironment in head and neck squamous cell carcinoma (HNSCC) [See @puram_single-cell_2017]  [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4675430.svg)](https://doi.org/10.5281/zenodo.4675430). More specifically, we will look at differential cell-cell communication patterns between tumors scoring high for a partial epithelial-mesenschymal transition (p-EMT) program vs low-scoring tumors.

In this vignette, we will prepare the data and analysis parameters, and then perform the MultiNicheNet analysis. 

The different steps of the MultiNicheNet analysis are the following:

* 1. Extract expression information from receiver and sender cell types: average expression for each gene per sample, and per group (normalized expression value and fraction of expressing cells with non-zero counts). 

* 2. Linking this expression information for ligands of the sender cell types to the corresponding receptors of the receiver cell types.

* 3. Perform genome-wide differential expression analysis of receiver and sender cell types to define DE genes between the conditions of interest. Based on this analysis, we can define the logFC/p-value of ligands in senders and receptors in receivers, and define the set of affected target genes in the receiver.

* 4. Predict NicheNet ligand activities and NicheNet ligand-target links based on these differential expression results

* 5. Use the information collected above to prioritize all sender-ligand --> receiver-receptor pairs.

In this vignette, we will demonstrate the use of a wrapper function to perform all these steps in one line of code. If you want to explore the different steps of MultiNicheNet one by one in more detail, you could check this other vignette: **[ADD THIS LATER]**.

After the MultiNicheNet analysis is done, we will explore the output of the analysis with different ways of visualization. 

# Step 0: Load required packages and NicheNet ligand-receptor network and ligand-target matrix

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(multinichenetr)
```

```{r}
# LR network
lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
lr_network = lr_network %>% filter(! database %in% c("ppi_prediction","ppi_prediction_go")) # We filter out these interactions because they are not really bona-fide LR pairs - NicheNet v2 prior models will be released relatively soon
lr_network = lr_network %>% dplyr::rename(ligand = from, receptor = to) %>% distinct(ligand, receptor)
```

```{r}
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
```

# Step 1: Prepare Seurat Objects for Sender and Receiver cells

This vignette demonstrates a multinichenet analysis with predefined sender and receiver cell types of interest.

In this case study, we want to study differences in cell-cell communication patterns between pEMT-high and pEMT-low tumors. The meta data columns that indicate the pEMT status of tumors are 'pEMT' and 'pEMT_fine', cell type is indicated in the 'celltype' column, and the sample is indicated by the 'tumor' column. 

Malignant cells as receiver

__User adaptation required__
```{r}
getwd()
seurat_obj_receiver = readRDS(url("https://zenodo.org/record/4675430/files/seurat_obj_hnscc.rds")) %>% subset(subset = celltype == "Malignant")
seurat_obj_receiver@meta.data$pEMT_fine = factor(seurat_obj_receiver@meta.data$pEMT_fine, levels = c("High","Medium","Low")) 
DimPlot(seurat_obj_receiver, group.by = "celltype")
DimPlot(seurat_obj_receiver, group.by = "tumor")
DimPlot(seurat_obj_receiver, group.by = "pEMT")
DimPlot(seurat_obj_receiver, group.by = "pEMT_fine")
```

Non-Malignant cells as sender

__User adaptation required__
```{r}
getwd()
seurat_obj_sender =readRDS(url("https://zenodo.org/record/4675430/files/seurat_obj_hnscc.rds")) %>% subset(subset = celltype != "Malignant")
seurat_obj_sender@meta.data$pEMT_fine = factor(seurat_obj_sender@meta.data$pEMT_fine, levels = c("High","Medium","Low")) 
DimPlot(seurat_obj_sender, group.by = "celltype")
DimPlot(seurat_obj_sender, group.by = "tumor")
DimPlot(seurat_obj_sender, group.by = "pEMT")
DimPlot(seurat_obj_sender, group.by = "pEMT_fine")
```

# Step 2: Prepare the cell-cell communication analysis

## Define senders and receivers of interest

```{r}
senders_oi = seurat_obj_sender %>% Idents() %>% unique()
receivers_oi = seurat_obj_receiver %>% Idents() %>% unique()

seurat_obj_sender = seurat_obj_sender %>% subset(idents = senders_oi)
seurat_obj_receiver = seurat_obj_receiver %>% subset(idents = receivers_oi)
```


We will now also check the number of cells per cell type condition combination, and the number of patients per condition.

__User adaptation required__
```{r}
table(seurat_obj_receiver@meta.data$celltype, seurat_obj_receiver@meta.data$tumor) # cell types vs samples
table(seurat_obj_receiver@meta.data$celltype, seurat_obj_receiver@meta.data$pEMT_fine) # cell types vs conditions
table(seurat_obj_receiver@meta.data$tumor, seurat_obj_receiver@meta.data$pEMT_fine) # samples vs conditions
```
__User adaptation required__
```{r}
table(seurat_obj_sender@meta.data$celltype, seurat_obj_sender@meta.data$tumor) # cell types vs samples
table(seurat_obj_sender@meta.data$celltype, seurat_obj_sender@meta.data$pEMT_fine) # cell types vs conditions
table(seurat_obj_sender@meta.data$tumor, seurat_obj_sender@meta.data$pEMT_fine) # samples vs conditions
```


As you can see, some Celltype-Sample combinations have 0 cells. It is possible that during DE analysis, some cell types will be removed from the analysis if there is not enough information to do a DE analysis.

# Step 2: Prepare the cell-cell communication analysis

## Define in which metadata columns we can find the **group**, **sample** and **cell type** IDs

For the group_id, we now choose for the 'pEMT_fine' column because we now want to compare the p-EMT-high vs the p-EMT-mid and p-EMT-low group

__User adaptation required__
```{r}
sample_id = "tumor"
group_id = "pEMT_fine"
celltype_id_receiver = "celltype"
celltype_id_sender = "celltype"
```

## Define the contrasts and covariates of interest for the DE analysis, and the minimal number of cells of a cell type that each sample should have to be considered for DE analysis of that cell type.

Here, we want to compare the p-EMT-high vs the p-EMT-mid and p-EMT-low group and find cell-cell communication events that are higher in high than mid and low pEMT. We don't have other covariates to correct for in this dataset.

Note the format to indicate the contrasts!

__User adaptation required__
```{r}
covariates = NA
contrasts_oi = c("'High-(Medium+Low)/2','Medium-(High+Low)/2','Low-(Medium+High)/2'") # no spaces between the different contrasts!
contrast_tbl = tibble(contrast = 
                        c("High-(Medium+Low)/2", "Medium-(High+Low)/2","Low-(Medium+High)/2"),  # division by 2 necessary because 2 groups to compare against !!
                      group = c("High","Medium","Low")) # division by 2 necessary because 2 groups to compare against !!
min_cells = 5 # here 5 for demonstration purposes - but recommended default is 10.

```

## Define the parameters for the NicheNet ligand activity analysis 

Here, we need to define the thresholds that will be used to consider genes as differentially expressed or not (logFC, p-value, decision whether to use adjusted or normal p-value, minimum fraction of cells that should express a gene in at least one sample in a group, whether to use the normal p-values or empirical null p-values). 

NicheNet ligand activity will then be calculated as the enrichment of predicted target genes of ligands in this set of DE genes compared to the genomic background. Here we choose for a minimum logFC of 0.50, maximum p-value of 0.05 (normal, not adjusted one, because of lack of statistical power due to pseudobulking and the fact that this dataset has only a few samples per group), and minimum fraction of expression of 0.05. 
For the NicheNet ligand-target inference, we also need to select which top n of the predicted target genes will be considered (here: top 250 targets per ligand).
We will also use the empirical p-values ([ADD REFERENCE]). This has mainly advantages when there would be some violations to the assumptions that should be fulfilled for having accurate theoretical p-values. In case all assumptions are fulfilled, using empirical p-values would not harm either.

__User adaptation recommended__
```{r}
logFC_threshold = 0.50
p_val_threshold = 0.05
fraction_cutoff = 0.05
p_val_adj = FALSE 
empirical_pval = TRUE
top_n_target = 250
```


## Define the weights of the prioritization of both expression, differential expression and NicheNet activity information

MultiNicheNet allows the user to define the weights of the following criteria to prioritize ligand-receptor interactions:

* Upregulation of the ligand in a sender cell type and/or upregulation of the receptor in a receiver cell type - in the condition of interest. : `scaled_lfc_ligand`, `scaled_p_val_ligand`, `scaled_lfc_receptor`, and `scaled_p_val_receptor`
* Sufficiently high expression levels of ligand and receptor in many samples of the same group (to mitigate the influence of outlier samples). : `fraction_expressing_ligand_receptor`
* Cell-type and condition specific expression of the ligand in the sender cell type and receptor in the receiver cell type (to mitigate the influence of upregulated but still relatively weakly expressed ligands/receptors) : `scaled_avg_exprs_ligand`,  `scaled_avg_frq_ligand`,  `scaled_avg_exprs_receptor`,  and `scaled_avg_frq_receptor`
* High NicheNet ligand activity, to further prioritize ligand-receptor pairs based on their predicted effect of the ligand-receptor interaction on the gene expression in the receiver cell type : `scaled_activity` and `scaled_activity_scaled` (scaled_activity=absolute value of ligand activity; scaled_activity_scaled=scaled ligand activity value that is comparable between different receiver settings - recommended to put more weight on this)
* High relative abundance of sender and/or receiver in the condition of interest: `scaled_abundance_sender` and `scaled_abundance_receiver`

We will set our preference for this dataset as follows:

__User adaptation recommended__
```{r}
prioritizing_weights_DE = c("scaled_lfc_ligand" = 1.5,
                         "scaled_p_val_ligand" = 1.5,
                         "scaled_lfc_receptor" = 1.5,
                         "scaled_p_val_receptor" = 1.5)
prioritizing_weights_activity = c("scaled_activity_scaled" = 2,
                                  "scaled_activity" = 2)

prioritizing_weights_expression_specificity = c("scaled_avg_exprs_ligand" = 1,
                         "scaled_avg_frq_ligand" = 1,
                         "scaled_avg_exprs_receptor" = 1,
                         "scaled_avg_frq_receptor" = 1)

prioritizing_weights_expression_sufficiency = c("fraction_expressing_ligand_receptor" = 2)

prioritizing_weights_relative_abundance = c( "scaled_abundance_sender" = 0,
                         "scaled_abundance_receiver" = 0)

```

```{r}
prioritizing_weights = c(prioritizing_weights_DE, 
                         prioritizing_weights_activity, 
                         prioritizing_weights_expression_specificity,
                         prioritizing_weights_expression_sufficiency, 
                         prioritizing_weights_relative_abundance)
```


# Step 3: Perform the cell-cell communication analysis

Now we will run the MultiNicheNet wrapper. In the function `multi_nichenet_analysis`, we need to specify that we use two Seurat objects (one with senders and one with receivers) by setting `sender_receiver_separate = TRUE`. This setting will call the underlying `multi_nichenet_analysis_separate` pipeline function.

To keep track of the different steps, we will here set `verbose = TRUE`

```{r}
output = multi_nichenet_analysis(seurat_obj_receiver = seurat_obj_receiver, seurat_obj_sender = seurat_obj_sender, 
                                celltype_id_receiver = celltype_id_receiver, celltype_id_sender = celltype_id_sender, sample_id = sample_id, group_id = group_id, 
                                lr_network = lr_network, ligand_target_matrix = ligand_target_matrix, contrasts_oi = contrasts_oi, contrast_tbl = contrast_tbl, covariates = covariates,
                                prioritizing_weights = prioritizing_weights, min_cells = min_cells, logFC_threshold = logFC_threshold, p_val_threshold = p_val_threshold,  
                                fraction_cutoff = fraction_cutoff, p_val_adj = p_val_adj, empirical_pval = empirical_pval, top_n_target = top_n_target, sender_receiver_separate = TRUE, verbose = TRUE)
```

The output of the MultiNicheNet analysis contains much information. We will now go over this step-by-step

## Check the returned tables in the output

### Average expression value and fraction of each cell type - sample combination

```{r}
output$receiver_info$avg_df %>% head()
output$receiver_info$frq_df %>% head()
output$receiver_info$avg_df_group %>% head()
output$receiver_info$frq_df_group %>% head()
output$receiver_info$rel_abundance_df %>% head()
```

```{r}
output$sender_info$avg_df %>% head()
output$sender_info$frq_df %>% head()
output$sender_info$avg_df_group %>% head()
output$sender_info$frq_df_group %>% head()
output$sender_info$rel_abundance_df %>% head()
```

### DE information for each cell type - contrast combination

```{r}
output$receiver_de %>% head()
```

```{r}
output$sender_de %>% head()
```

### Output of the NicheNet ligand activity analysis, and the NicheNet ligand-target inference

```{r}
output$ligand_activities_targets_DEgenes$ligand_activities %>% head()
```

```{r}
output$ligand_activities_targets_DEgenes$de_genes_df %>% head()
```

### Tables with the final prioritization scores (results per group and per sample)

```{r}
output$prioritization_tables$group_prioritization_tbl %>% head()
```

```{r}
output$prioritization_tables$sample_prioritization_tbl %>% head()
```

Based on these prioritization tables, we will define which interactions to visualize in the different plots below.

# Step 4: Visualize the results of the cell-cell communication analysis

In a first instance, we will look at the broad overview of prioritized interactions via condition-specific Circos plots.
We will filter on the prioritization score after removing interactions that are not upregulated in the condition of interest, and that are in no patient expressed in more than 5% of cells of a cell type (cf `fraction_cutoff`).

We will look here at the top100 predictions across all contrasts (high vs low; and low vs high), senders, and receivers of interest.

## Circos plot of top-prioritized links

(remark: there are still some bugs in the automatic circos plot function)

```{r, fig.width=15, fig.height=12}
prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(fraction_ligand_group > fraction_cutoff & fraction_receptor_group > fraction_cutoff) %>% 
  distinct(id, sender, receiver, ligand, receptor, group, prioritization_score, ligand_receptor_lfc_avg, fraction_expressing_ligand_receptor) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% top_n(100, prioritization_score) 

prioritized_tbl_oi %>% group_by(group) %>% count()

prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(id %in% prioritized_tbl_oi$id) %>% 
  distinct(id, sender, receiver, ligand, receptor, group) %>% left_join(prioritized_tbl_oi)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0


n_senders = prioritized_tbl_oi$sender %>% unique() %>% length()
n_receivers = prioritized_tbl_oi$receiver %>% unique() %>% length()

  
colors_sender = c("red", "royalblue") %>% magrittr::set_names(prioritized_tbl_oi$sender %>% unique())
colors_receiver = c("orange") %>% magrittr::set_names(prioritized_tbl_oi$receiver %>% unique())

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)

```

## Visualization of scaled_LR_prod per sample

Now we will visualize per sample the scaled product of ligand and receptor expression. Samples that were left out of the DE analysis are indicated with a smaller dot (this helps to indicate the samples that did not contribute to the calculation of the logFC, and thus not contributed to the final prioritization)

We will now check the top75 interactions specific for the pEMT high group

```{r}
group_oi = "High"
```

```{r, fig.height=18, fig.width=12}

output$prioritization_tables$sample_prioritization_tbl$group = factor(output$prioritization_tables$sample_prioritization_tbl$group, levels = c("High","Medium","Low"))

prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  distinct(id, sender, receiver, lr_interaction, group, ligand_receptor_lfc_avg, activity_scaled, fraction_expressing_ligand_receptor,  prioritization_score) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% 
  filter(group == group_oi) %>% group_by(group) %>% top_n(75, prioritization_score) 

plot_oi = make_sample_lr_prod_plots(output$prioritization_tables, prioritized_tbl_oi)
plot_oi
```


Next to these LR expression products, we can also plot the NicheNet ligand activities of the ligand in the receiver.

```{r, fig.height=18, fig.width=18}
output$prioritization_tables$group_prioritization_tbl$group = factor(output$prioritization_tables$group_prioritization_tbl$group, levels = c("High","Medium","Low"))

prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  distinct(id, sender, receiver, lr_interaction, group, ligand_receptor_lfc_avg, activity_scaled, fraction_expressing_ligand_receptor,  prioritization_score) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% 
  filter(group == group_oi) %>% group_by(group) %>% top_n(75, prioritization_score) 

plot_oi = make_sample_lr_prod_activity_plots(output$prioritization_tables, prioritized_tbl_oi, widths = c(5,1,1))
plot_oi
```

## Visualization of expression-logFC per group and ligand activity

Next type of plots will show the logFC of LR pairs across all sender-receiver pairs that are selected, and add the ligand activity next to it.

```{r}
receiver_oi = "Malignant"
group_oi = "High"
```

```{r, fig.width=18, fig.height=9}

prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(fraction_ligand_group > fraction_cutoff & fraction_receptor_group > fraction_cutoff) %>% 
  distinct(id, sender, receiver, lr_interaction, group, ligand_receptor_lfc_avg, activity_scaled, fraction_ligand_group, fraction_expressing_ligand_receptor, scaled_avg_exprs_ligand, prioritization_score) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% 
  filter(group == group_oi & receiver == receiver_oi) %>% top_n(75, prioritization_score) 

plot_oi = make_group_lfc_exprs_activity_plot(output$prioritization_tables, prioritized_tbl_oi, receiver_oi, heights = c(5,1,1))
plot_oi

```

## Visualization of ligand-activity, ligand-target links, and target gene expression

In another type of plot, we can visualize the ligand activities for a group-receiver combination, and show the predicted ligand-target links, and also the expression of the predicted target genes across samples.

First: show this for a selection of ligands with high ligand activities:

```{r}
group_oi = "High"
receiver_oi = "Malignant"
prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(fraction_ligand_group > fraction_cutoff & fraction_receptor_group > fraction_cutoff) %>% 
  distinct(id, sender, receiver, ligand, receptor, group, prioritization_score, ligand_receptor_lfc_avg, fraction_expressing_ligand_receptor, activity_scaled) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% 
  filter(group == group_oi & receiver == receiver_oi) %>% 
  group_by(group) %>% top_n(50, prioritization_score) %>% top_n(25, activity_scaled) %>% arrange(-activity_scaled)
```

```{r, fig.width=18, fig.height=10}
combined_plot = make_ligand_activity_target_plot(group_oi, receiver_oi, prioritized_tbl_oi, output$ligand_activities_targets_DEgenes, contrast_tbl, output$grouping_tbl, output$receiver_info, plot_legend = FALSE)
combined_plot
```

Now: show this for a selection of ligands with high general prioritization scores, not necessarily high ligand activities.

```{r}
group_oi = "High"
receiver_oi = "Malignant"
prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(fraction_ligand_group > fraction_cutoff & fraction_receptor_group > fraction_cutoff) %>% 
  distinct(id, sender, receiver, ligand, receptor, group, prioritization_score, ligand_receptor_lfc_avg, fraction_expressing_ligand_receptor, activity_scaled) %>% 
  filter(ligand_receptor_lfc_avg > 0 & fraction_expressing_ligand_receptor > 0) %>% 
  filter(group == group_oi & receiver == receiver_oi) %>% 
  group_by(group) %>% top_n(25, prioritization_score) %>% arrange(-activity_scaled)
```

```{r, fig.width=18, fig.height=10}
combined_plot = make_ligand_activity_target_plot(group_oi, receiver_oi, prioritized_tbl_oi, output$ligand_activities_targets_DEgenes, contrast_tbl, output$grouping_tbl, output$receiver_info, plot_legend = FALSE)
combined_plot
```

## Show ligand activities for each receiver-group combination

In the next type of plot, we plot all the ligand activities (Both scaled and absolute activities) of each receiver-group combination. This can give us some insights in active signaling pathways across groups. Note that we can thus show top ligands based on ligand activity - agnostic of expression in sender. 

```{r, fig.width=5, fig.height=8}

ligands_oi = output$prioritization_tables$ligand_activities_target_de_tbl %>% inner_join(contrast_tbl) %>% 
  group_by(group, receiver) %>% distinct(ligand, receiver, group, activity) %>% 
  top_n(5, activity) %>% pull(ligand) %>% unique()

plot_oi = make_ligand_activity_plots(output$prioritization_tables, ligands_oi, contrast_tbl, widths = NULL)
plot_oi

```

Or we can do this plot for ligands, while considering the general priorization score (which considers expression information etc)

Show top ligands based on prioritization scores

```{r, fig.width=5, fig.height=8}

ligands_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  group_by(group, receiver) %>% distinct(ligand, receiver, group, prioritization_score) %>% 
  top_n(5, prioritization_score) %>% pull(ligand) %>% unique()

plot_oi = make_ligand_activity_plots(output$prioritization_tables, ligands_oi, contrast_tbl, widths = NULL)
plot_oi

```

## Zoom in on specific ligand-receptor interactions: show their expression in the single-cell data!

Single-cell-based Nebulosa, Feature, and Violin plots of ligand-receptor interaction of interest: `make_ligand_receptor_nebulosa_feature_plot` and `make_ligand_receptor_violin_plot`

It is often useful to zoom in on specific ligand-receptor interactions of interest by looking in more detail to their expression at the single cell level 

Check the highest scoring links based on the general prioritization score. Here we will pick one of those to visualize.

```{r}
prioritized_tbl_oi = output$prioritization_tables$group_prioritization_tbl %>% 
  filter(fraction_ligand_group > fraction_cutoff & fraction_receptor_group > fraction_cutoff) %>% 
  distinct(id, sender, receiver, ligand, receptor, group, prioritization_score) %>% 
  group_by(group, receiver) %>% top_n(5, prioritization_score) 
prioritized_tbl_oi
```

```{r}
ligand_oi = "IL24"
receptor_oi = "IL20RB"
group_oi = "High"
sender_oi = "CAF"
receiver_oi = "Malignant"
```

Nebulosa and Feature plot of the ligand in the sender cell type and the receptor in the receiver cell type (split per condition)

```{r, fig.width=15, fig.height=15}
plot_list = make_ligand_receptor_nebulosa_feature_plot(seurat_obj_sender = seurat_obj_sender, seurat_obj_receiver = seurat_obj_receiver, ligand_oi = ligand_oi, receptor_oi = receptor_oi, group_oi = group_oi, group_id = group_id, celltype_id_sender = celltype_id_sender, celltype_id_receiver = celltype_id_receiver, senders_oi = c("myofibroblast","CAF"), receivers_oi = c("Malignant"), prioritized_tbl_oi = prioritized_tbl_oi)
plot_list$nebulosa # not recommended for Smart-seq data
plot_list$feature
```
Pooled single-cell and sample-specific single-cell violin plots of ligand and receptor expression in respectively sender and receiver.


```{r, fig.width=13, fig.height=8}
plot_list2 = make_ligand_receptor_violin_plot(seurat_obj_sender = seurat_obj_sender, seurat_obj_receiver = seurat_obj_receiver, ligand_oi = ligand_oi, receptor_oi = receptor_oi, group_oi = group_oi, group_id = group_id, sender_oi = sender_oi, receiver_oi = receiver_oi, sample_id = sample_id, celltype_id_sender = celltype_id_sender, celltype_id_receiver = celltype_id_receiver, prioritized_tbl_oi = prioritized_tbl_oi)
plot_list2$violin_group
plot_list2$violin_sample
```

## Zoom in on specific ligand-target interactions: show their expression in the single-cell data!

Make target gene violin and nebulosa plots: `make_target_violin_plot` and `make_target_nebulosa_feature_plot`

```{r}
receiver_oi = "Malignant"
group_oi = "High"

output$ligand_activities_targets_DEgenes$de_genes_df %>% inner_join(contrast_tbl) %>% filter(p_val <= 0.05) %>% filter(group == group_oi) %>% filter(receiver == receiver_oi) %>% arrange(p_val) %>% top_n(100, logFC)  
```

RAB31: interesting gene  

```{r, fig.width=18, fig.height=8}
target_oi = "NEFM"

make_target_violin_plot(seurat_obj_receiver = seurat_obj_receiver, target_oi = target_oi, receiver_oi = receiver_oi, group_oi = group_oi, group_id = group_id, sample_id, celltype_id_receiver = celltype_id_receiver, output$prioritization_tables$group_prioritization_tbl)
make_target_nebulosa_feature_plot(seurat_obj_receiver = seurat_obj_receiver, target_oi = target_oi, group_oi = group_oi, group_id = group_id, celltype_id_receiver = celltype_id_receiver, receivers_oi = c("Malignant"), output$prioritization_tables$group_prioritization_tbl) 

```

## Make Dotplot for all DE genes/targets

Note: DE here determined based on the parameters used for the MultiNicheNet analysis (cf above): this means that DE genes are here not based on the p-value corrected for multiple testing!

```{r}
receiver_oi = "Malignant"
group_oi = "High"

targets_oi = output$ligand_activities_targets_DEgenes$de_genes_df %>% inner_join(contrast_tbl) %>% filter(group == group_oi) %>% arrange(p_val)  %>% filter(receiver == receiver_oi) %>% top_n(200, logFC) %>% pull(gene) %>% unique() 

```

```{r, fig.width=8, fig.height=30}
p_target = make_sample_target_plots(receiver_info = output$receiver_info, targets_oi, receiver_oi, output$grouping_tbl)
p_target + ggtitle(paste0("DE genes in ",group_oi, " in celltype ",receiver_oi))
```

## References


